Distancing alters the controllability of emotional states by affecting both intrinsic stability and extrinsic sensitivity

Version 1

Abstract

Abstract

Background

Emotion regulation strategies such as distancing are a core component of many evidence-based, effective psychotherapeutic interventions. They allow individuals to exert more ‘control’ over their emotional state. However, objectively disentangling how emotion regulation increases control has been difficult for reasons including a lack of a coherent theoretical framework for emotion control and insufficient experimental control over external inputs. Here, we apply a well-established theoretical framework for controllability to a tightly controlled experimental setup to examine the computational mechanisms by which emotion regulation interventions enhance emotional controllability.

Methods

109 participants were randomized to either a short emotion regulation intervention (distancing) or a control intervention. Both before and after the intervention, participants reported their emotional state along five dimensions repeatedly while watching a series of short, standardized, emotional video clips. A Kalman Filter was used to quantify how multidimensional emotional states changed with video inputs. The consequences of the emotion regulation intervention were examined by Bayesian model comparison, comparing models allowing for a change in intrinsic dynamics and/or input weights. Controllability was quantified using the controllability Gramian.

Results

The Kalman filter captured participants’ emotional trajectories, showing that emotional states were affected by the emotional videos; persisted; and interacted with each other. The distancing strategy made emotional states less externally controllable. It did so by altering two aspects of the dynamical system: by stabilizing specific emotional patterns and by reducing the impact of the external video clips.

Conclusions

Our study used a novel approach to examine emotion regulation, finding that a brief distancing intervention increased perceived emotion control by reducing how much external stimuli can control emotional states. This is due to both an increase in the intrinsic stability of certain emotional states; and a reduction of the sensitivity to certain extrinsic affective stimuli.

1 Introduction

Psychotherapeutic interventions are effective treatments for depression and anxiety, but the mechanisms of action responsible for these effects remain poorly understood (Kazdin, 2007; Carey et al., 2020). In addition, there is considerable variation between individuals regarding treatment response (Goldfried, 2013). Better targeting psychotherapeutic interventions according to individual needs or characteristics might improve therapy effectiveness. This requires understanding how existing psychological treatments work and for whom they work.

Previous research studying the mechanisms underlying psychological treatments has mostly focused on changes after the complete treatment courses (Klug et al., 2012). However, psychotherapeutic treatments are complex interventions involving different components (Luborsky et al., 2002), likely affecting different behavioural processes and acting via different mechanisms. As such, changes observed after the full course of treatment reflect the many many aspects of the intervention. In contrast, little research exists on how different components engage specific mechanisms.

In this study, we build on recent suggestions (Reiter et al., 2021; Wolpert et al., 2021; Holmes et al., 2018) and a promising new research direction (Brown et al., 2021; Huys et al., 2022; Dercon et al., 2023; Norbury et al., 2023), using cognitive computational techniques to study isolated interventions. This approach aims to better characterize and understand the change mechanisms of specific, key psychotherapeutic interventions. Specifically, we examine the effect of distancing—an emotion regulation technique—on the dynamics of emotions.

Emotion regulation strategies are an essential element of many psychotherapeutic treatments (Gross, 1998). Some emotion regulation strategies can alleviate symptoms of mood disorders and generally improve well-being (Berking et al., 2013; Boemo et al., 2022; Powers and LaBar, 2019; Somerville et al., 2022). Here, we define emotion regulation as using explicit strategies to intentionally up- or down-regulate positive or negative emotions. Effective strategies include problem-solving, reappraisal, acceptance, and distancing techniques (McRae and Gross, 2020; Webb et al., 2012). Distancing involves simulating a new perspective to increase the psychological distance from an event or situation and, with that, the emotional impact of a stimulus (Powers and LaBar, 2019; Webb et al., 2012). Distancing has been shown to reduce selfreported emotional experience reliably (Koenigsberg et al., 2009; Vrticka et al., 2012; Winecoff et al., 2011, 2013) and is associated with decreased amygdala activity even beyond the period of active regulation (Eippert et al., 2007; Walter et al., 2009; Domes et al., 2010). Distancing techniques are practical because they can be implemented in various situations with relatively low attentional demands and behavioural disruption (Powers and LaBar, 2019). We focus on distancing to gain insights into the mechanism of this specific treatment component in the hope of better understanding some of the complex processes underlying psychotherapeutic treatment effects.

Furthermore, emotion regulation research has mostly focused on the effects on individual emotions. Different emotions, however, are often related and influence each other. For instance, sadness can increase the likelihood of experiencing anger while emotions of different valence or arousal tend to inhibit each other: inducing happiness reduces experienced sadness, yet mixed states with both sadness and happiness can also occur. These interactions induce temporal dependencies between emotions, meaning that emotional states as a whole form a dynamical system fluctuating over time (Malamud and Huys, 2024; Durstewitz et al., 2020; Kuppens and Verduyn, 2017; Lange et al., 2022). As such, emotion regulation may be best conceptualized as the regulation of a dynamical system of emotions rather than as involving the regulation of individual emotions in isolation.

In recent years, network models have been developed to account for the dynamical systems properties of emotions (Borsboom and Cramer, 2013; Bos et al., 2017; Bringmann et al., 2013; Epskamp et al., 2018). This work has identified individual differences in affect dynamics which are linked to mood disorders (Bringmann et al., 2016; Kuppens et al., 2012; Pe et al., 2015, 2016; Sperry et al., 2020; Trull et al., 2015; Leemput et al., 2014). For example, increased inertia (temporal autocorrelation or how well an emotion can maintain itself) of negative affect has been identified in people with depression (Brose et al., 2015; Houben et al., 2015; Kuppens et al., 2010; Koval et al., 2012, 2013). Most interestingly, it has been suggested that there are discrete stable states (e.g. a depressed vs a happy state), so-called attractor states, between which people can transition (Durstewitz et al., 2020; Leemput et al., 2014; Wichers et al., 2016; Kuppens et al., 2012; Hosenfeld et al., 2015). Within an attractor state, the system is usually resistant to change, and under small disturbances (e.g. a stressful week), a person converges back to the current attractor state. A transition to another attractor state may occur if a perturbation is large enough or accumulated over time (Nelson et al., 2017). The transitions between states in a dynamical system are influenced by two important factors: the system’s intrinsic characteristics and its sensitivity to external driving forces.

Past work mostly focused on dynamical properties inferred from the emotion self-reports alone and has often neglected the role of external factors (Boemo et al., 2022; Malamud and Huys, 2024). This is a critical omission as the dynamical properties of a system are not fully identifiable unless the inputs are known — inputs can nearly arbitrarily alter the apparent dynamical system. For instance, persistent sad mood could be due to a constant external stimulus producing sadness rather than due to an internal persistence of sad mood. Ignoring the immediate context within which emotions fluctuate may hence lead to wrong conclusions about the underlying affective system. Indeed, it is well known that external stimuli profoundly impact the dynamics of emotional states, particularly in laboratory settings (Asutay et al., 2022; Rutledge et al., 2014; Villano et al., 2020; Vanhasbroeck et al., 2022).

Since both the natural changes in emotions over time and the impact of external factors on emotions are important, there is a compelling rationale to explore the potential effects of emotion regulation on these two key aspects of emotion processes. By doing so, we can gain valuable insights into the mechanisms underlying emotional regulation and better understand how emotions are controlled. Understanding which emotions can be most effectively changed through a certain techniques can provide clearer insights into the targeted effects of such interventions. Additionally, identifying reliable indicators of whether the emotional system is resistant or adaptable to changes could open up possibilities for creating personalized interventions.

Therefore, our aim was to examine the dynamic network effect of emotion regulation, specifically by considering an individual’s immediate context. We positioned emotion regulation within the framework of dynamical system theory and utilized tools from control theory to characterize the overall control properties of the emotional system. To briefly explain, a dynamical system is a series of linked differential equations (Brunton and Kutz, 2019; Durstewitz et al., 2020), each describing how one variable (here an emotion) changes over time. It permits emotions to influence each other and exhibits rich dynamical properties. The equations can incorporate the concept of control by allowing an input to drive one or more variables. Controllability, in this context, is a characteristic that depends on both how sensitive the system is to inputs and its inherent dynamics. It reflects how easy or hard it is to drive the system towards certain states. For instance, if all emotions are positively linked to sad mood, then a happy state would be challenging to achieve, reducing the system’s controllability of single emotions.

In this study, we investigated how external emotionally-charged inputs influenced self-reported emotions over time (across multiple dimensions), and whether these inputs were essential to explaining the evolution of a rich, multidimensional emotional state. This enabled us to formally examine the impact of a distancing intervention on emotion dynamics whilst disentangling effects on intrinsic dynamics from alterations of external inputs. We also examined whether intervention-induced changes were moderated by measures of depressive or anxiety symptoms or difficulties in regulating emotions.

2 Methods

2.1 Participants

109 participants 18 years or older with current UK residence were recruited online on Prolific Academic (www.prolific.co). Prolific is widely used and reliable platform for online research due to its diverse participant pool and robust quality control measures (Peer et al., 2017; Palan and Schitter, 2018).

An a priori power analysis based on data from a pilot study estimated a minimum sample size of N = 109 (cf. Supplementary Materials D Power Analysis).

2.2 Procedure and Task

After indicating interest in the online recruitment platform, participants were forwarded to an electronic form of the participant information sheet. They could then provide electronic consent through an online form before being redirected to the experiment. The study duration was approximately 45 minutes, and participants were reimbursed £7.50/h through Prolific Academic after completion of study procedures.

In the experiment (Fig. 1), participants saw a sequence of short emotional video clips, each lasting 2–10 seconds and chosen for their efficacy in eliciting certain emotions. Video clips were from a previously-published database, with validated emotion ratings across multiple categories (cf. 2.2.2 Emotion-Inducing Stimuli; Cowen and Keltner 2017). After each video clip, participants reported their current emotional state in terms of two positive (amused and calm) and three negative emotions (disgusted, anxious, and sad). Emotions of disgust, anxiousness, and amusement were chosen as they were thought to be highly sensitive to video clip inputs. Conversely, sad and calm appear to be more persistent emotions which we expected to be informative about stability. Participants were instructed to use a slider to indicate how strongly they felt each of the emotions at that moment in time, with options ranging from “not at all” to “very”. Participants had 30 seconds to report their emotional state. If they did not manage to rate all emotions within 30 seconds, the experiment moved on to the next video without their completed rating.

After watching the first block of 54 video clips, participants were randomized to undergo either a distancing (emotion regulation) or a relaxation (control) intervention (cf. 2.2.1 Intervention). Following the intervention, participants watched a second block of 54 video clips and again rated their emotions after each video. A different set of video clips was shown, but the sequence of emotions targeted was matched to the first block.

Task Description

The large arrow at the top displays the course of the experiment. Both video blocks comprised 54 video clips (2–10 seconds), and after each video clip, participants had 30 seconds to rate their emotional experience based on five emotions. Both blocks also had a baseline emotion rating before the videos started. Participants underwent either a distancing (emotion regulation) or a relaxation (control) intervention between the two video blocks. After the second block of video clips, participants completed three standardized psychological questionnaires measuring symptoms of depression (PHQ-9) and anxiety (GAD-7) and emotion regulation difficulty (DERS-18). PHQ-9: Patient Health Questionnaire. GAD-7: Generalized Anxiety Disorder Assessment. DERS-18: Difficulties in Emotion Regulation Scale.

2.2.1 Intervention

The emotion regulation intervention was based on a distancing appraisal strategy, i.e. “Leaves on a Stream” (adapted from Hayes et al. 2006). This technique involves viewing emotions and thoughts as events passing through one’s mind rather than getting “sucked in” by them. The script employed a visualization strategy, instructing participants to imagine they were standing by a stream with leaves floating gently past them. They were then told: “When an emotion or thought comes up, imagine you place the thought on one of those leaves and that you are watching the leaf - carrying your emotion or thought - float away, disappearing behind a corner or in the distance.” In the control intervention, participants were asked to engage in a relaxation exercise. They were told the same storyline as in the distancing intervention but without connecting the stream and leaves to their emotions and thoughts (the full text of both interventions is reproduced in the Supplementary Materials A Intervention Text).

Furthermore, in the group allocated to the distancing intervention, the phrasing of the question about their feelings in the second video block reminded participants about the intervention, stating: “You observed your emotions and let them pass like the leaves floating by on the stream.” This was not the case for the control group.

To reduce demand effects, the instructions were framed to suggest that the experiment was attempting to understand whether the distancing was helpful or not, i.e. participants were explicitly encouraged to report that the intervention was not helpful if they did not perceive it as effective.

2.2.2 Emotion-Inducing Stimuli

The emotion eliciting video clips stem from a validated database (Cowen and Keltner, 2017) (https://s3-us-west-1.amazonaws.com/emogifs/uncensored.html). This study analyzed emotional responses to 2,185 short videos, identifying 27 distinct emotional experiences categorized through a survey completed by 853 English-speaking US participants. The videos were collected from online sources depicting diverse emotionally significant situations, including cute animals, natural landscapes, distressing scenes such as feces and vomit, accidents and dangerous stunts, and many others.

For our experiment, we selected video clips that fit into five target emotion categories: Amusement/Joy, Disgust/Horror, Sadness/Sympathy, Calmness/Aesthetic Adoration, and Anxiety/Fear. We chose clips that showed a high mean and a low entropy in the ratings from Cowen and Keltner (2017) for these emotion categories, to ensure they would specifically evoke the intended emotion. We identified 20 videos from each of the five categories, resulting in a total of 100 videos. The ratings from Cowen and Keltner (2017) for each chosen video and the mean ratings over videos from an emotion category are shown in Supplementary Materials B Video Clips Fig. B.2, respectively Fig. B.1.

2.2.3 Study Sequence

In total, each participant viewed 54 video clips before and 54 video clips after the intervention. The video order was pseudorandomized for the first half. In addition, two randomly chosen videos were repeated three times within the first block to investigate the reliability of the ratings. The second block of videos contained different video clips, but the sequence of target emotions was the same as in the first block (including the repeated videos). All participants saw the identical video clip sequences, but before and after intervention sequences were counterbalanced across participants. Participants provided 110 ratings in total (55 before and 55 after the intervention) including a baseline rating before the first video of each block. The length of the experiment was based on simulation and recovery work (cf. Supplementary Materials C Experimental Design).

2.2.4 Self-Report Measures

At the end, participants were asked to complete three self-report questionnaires. To assess depressive and anxiety symptoms, we used the Patient Health Questionnaire (PHQ-9; Kroenke et al. 2001) and the Generalized Anxiety Disorder Assessment (GAD-7; Spitzer et al. 2006), respectively. In addition, we used a short version of the Difficulties in Emotion Regulation Scale (DERS-18; Victor and Klonsky 2016) to assess participants’ ability to identify, accept, and manage their emotional experiences. The questionnaires were conducted after the experiment to prevent them from affecting participants’ behaviour during the study.

2.2.5 Attention Checks

To maintain and monitor attention, participants were asked to detect a black cross, which could be shown briefly before the video clip started. 10 such attention checks were included, 4 in the first block and 6 in the second. Participants with an accuracy of less than 70% on the attention checks were excluded.

2.3 Computational Modelling

During the experiment, we asked participants to report their emotions after watching each video, resulting in a sequence of emotional state reports. To capture the temporal structure of this data, we used a linear Gaussian state space model, specifically the time-invariant Kalman Filter (Durbin and Koopman, 2012; Roweis and Ghahramani, 1999). This model is simple and tractable, accounting for possible errors in participants’ self-reports and allows for external inputs that might influence their emotions. Essentially, this model helps us track the true emotional process as it evolves based on the participants’ reports.

The Kalman Filter suggests an underlying latent process, referred to as the true emotional process, evolving in response to a sequence of emotional state reports xtt = 1T conditional on video inputs utt = 1T (cf. to Fig. 2A).

The latent (unobserved) emotional state vector zt comprises the activation of each emotion category zti at time t. These latent emotions are assumed to evolve according to a discrete, linear first-order Markov process. The matrix A defines the dynamics matrix, which captures the degree to which an emotion carries over from one time point to the next and the degree to which it predicts future emotions. The matrix C captures the impact of videos, with the entry Cij capturing the impact of a video of category j on emotion category i. The video category shown at each time point was identified through one-hot labelling in the binary input vector ut, i.e. each video was identified in the vector u by the main emotion it was targeting. The subjects’ ratings xi on each category i were mapped directly onto the latent emotion states zi. Critically, however, Gaussian noise was added to both the latent temporal evolution and the observation processes with separate diagonal covariances Σ and Γ, respectively. This is one of the key differences to standard emotional time-series analyses (Borsboom et al., 2021), in that this explicitly allows for noise or errors in the emotion ratings, and this noise can be ‘smoothed over’ through the filtering process.

2.3.1 Model Comparison

To understand the dynamics of the emotion reports elicited by the video sequences, we asked whether the different components of the Kalman Filter were indeed necessary to provide a parsimonious account of the observed data. To do this, we built increasingly complex models. The simplest model included only Gaussian noise, i.e. it assumed emotion ratings varied randomly over time. The next model additionally contained either a dynamics matrix (A), an input weight matrix (C) or both. The most complex model included a dynamics matrix (A), an input weight matrix (C) and a bias (h). Additionally, we examined variations of these models where we constrained the input matrix C to be diagonal. Finally, models were fitted separately to each individual’s emotion time-series using the python package Pykalman (https://pykalman.github.io/).

We calculated the Bayesian Information Criterion (BIC; Schwarz 1978) for each individual for each model based on an individual’s model loglikelihood by penalizing for the number of parameters in the model. Models were then compared using the BIC at the group level. Cumulative model weights for the most parsimonious model are computed as a proportion of the total amount of predictive power provided by the full set of models contained in the model being assessed: ωj=e−ΔBICj∑i=1Ne−ΔBICi where ΔBICi = BICi — min(BIC), j indicates the most parsimonious model and N the number of models.

2.3.2 Stability

After extracting the parameters of the most parsimonious model, we investigated the eigenstruc-ture of the dynamical system. Briefly, a linear dynamical system where variables interact, like the Kalman Filter model, can be decomposed into separate systems of non-interacting variables. This is achieved through an eigendecomposition of the dynamics matrix. By projecting the vector of state variables z on each of the eigenvectors, new combined state variables z˜ (eigenmodes) can be defined, which evolve independently, i.e. no longer interact. Hence, these new combined state variables identify the effective emotional combinations which determine the evolution of an individual’s affective state. The eigenvector corresponding to the largest eigenvalue of the dynamics matrix A identifies the most stable combination of emotions. In contrast, the eigenvector corresponding to the smallest eigenvalue identifies the combination of emotions that is most transitory and least persistent (for an example cf. Fig. 2B-D).

2.3.3 Controllability

Next, we investigated the controllability of the dynamical system. A system is more controllable if smaller inputs u are required to move its state z to any required value. We computed the controllability Gramian (C) for each participant as follows: C=[ CACA2C…An−1C ] (3)

The controllability Gramian (C) combines the dynamics matrix (A) and the weights of the external input (C) to the dynamical system. If the rank of this controllability matrix is equal to the system’s dimension, the system is controllable. A controllable system means that any state z can be achieved through the appropriate choice of external inputs u. How controllable the system is captured by the strength of input |u| required. We investigated the characteristics of the controllability Gramian using singular value decomposition. Unitary vectors of the controllability Gramian define an energy ellipsoid (Fig. 2E). Unitary vectors corresponding to higher singular values identify the more controllable directions in the state space and vice versa. The more controllable a direction is, the less input energy is required to steer the system in that specific direction. In other words, an input of a given strength |u| can move the system further in a direction which aligns with a more controllable direction than a less controllable one.

Dynamical System and Controllability

A) shows a graph visualization of the linear dynamical model, including external inputs (ut; emotional video clips). zt describes the latent (unobserved) emotion states evolving based on a Markov process and directly mapping onto the acquired emotion ratings xt. B) shows the trajectories of a two-dimensional system (ratings of “amused” and “anxious”) starting from a randomly chosen initial point without external inputs. Whereas anxiety decays independently of amusement, amusement is influenced by anxiety and thus, the trajectory of amusement is more complex and does not simply exponentially decay to zero. However, the more anxiety decays, the more the influence of anxiety on amusement decreases; and both variables converge towards zero. C) shows the trajectory of both emotion ratings plotted against each other. The blue x indicates the starting point. D) displays the independently evolving trajectories of the transformed variables z˜ resulting from the projection of the state variables z onto the eigenvectors of the dynamics matrix A. E) The unitary vectors (V) of the controllability matrix define an energy ellipsoid where the unitary directions corresponding to higher singular values (S) are more controllable and vice versa. That means with the same effort one can go further into the most controllable direction (v1) and least far into the least controllable direction (v3)

2.4 Statistical Analysis

We used one-sided two-sample t-tests to test whether the emotion ratings averaged over video clips from the same video category for the emotion which was aimed to be elicited were higher than for the other emotions. In addition, two-sided one-sample t-tests were performed to test whether the mean emotion ratings, autocorrelation coefficients, and cross-correlation coefficients of emotion time-series significantly differed from zero. Those tests were all conducted on data before the intervention.

To investigate intervention effects, the principal analyses was a two-sample Hotelling T2 tests to compare multivariate variables (e.g. eigendirections) between groups after the intervention. The randomized group allocation allowed us to focus on potential effects after the intervention. We also performed multivariate ANOVAs (MANOVAs) to test for an interaction effect between time (before and after intervention) and group. To zoom into single emotions, we used non-parametric tests because most dynamical and controllability features were not normally distributed. Mann-Whitney U tests were conducted to compare emotion variables, such as mean emotion ratings and eigenvector directions, between the intervention groups and one-sample Wilcoxon signed-rank test to compare variables before and after the intervention within an intervention group.

Multiple linear regressions were performed to investigate associations between emotion dynamics (dependent variable; DV) and symptoms (independent variable; IV) controlling for the intervention group (G): DV = β0 + β1IV + β2G. Finally, to investigate whether psychological well-being moderated the effect of the intervention on emotion ratings, we examined the interaction effect between symptom score and intervention group: DV = β0 + β1IV + β2G. All variables were z-scored for the regressions. For all above-mentioned analysis types, we used Bonferroni-correction to correct for the number of conducted tests.

3 Results

3.1 Participants

109 participants completed the experiment, but one participant was excluded due to attention check failure. Analyses hence focused on, 108 participants (57 randomized to the distancing intervention).

3.2 Eliciting Complex Emotional States With Videos

In the first video block before the intervention, each video induced the dominant (targeted) emotion as intended (Fig. 3A). Changes in the dominant emotion ratings for each video were significantly higher than changes in other emotions (t ∈ [5.9, 26.2], all p < 0.001; Fig. 3B and Supplementary Material Table I.2). This shows that the sequence of emotional videos reliably induced the targeted emotions. The correlation between our participants’ emotion ratings and those reported by Cowen and Keltner (2017) was moderate to strong across all emotion categories, with correlations of r = 0.74 for Disgust/Horror, r = 0.65 for Amusement/Joy, r = 0.5 for Calmness/Aesthetic Adoration, r = 0.71 for Anxiety/Fear, and r = 0.6 for Sadness/Sympathy (all p < 0.001). Thus, we replicated the ratings from Cowen and Keltner (2017) reasonably well.

Furthermore, changes in emotions other than the dominant one within a video category were significantly different from zero (|t| ∈ [5.1, 19.3], all p < 0.001; Fig. 3B and Supplementary Material Table I.2). This indicates that the videos induced complex and multi-faceted emotional states. We also observed that emotional responses had a dynamic component. This was evident in the autocorrelation of each emotion rating, where an effect of the previous time-point is apparent during the rating of the next video (AR(1) ∈ [0.07 - 0.25], all p ≤ 0.001; cf. Fig. 3C and Supplementary Material Table I.3).

Emotion Ratings

A) The heatmap shows for each emotion eliciting video category the ratings from Cowen and Keltner (2017) averaged over the two emotion categories of interest (left y-axis). The coloured lines with dots report the emotion ratings from our experiment for each emotion after watching a video clip averaged over participants before the intervention occurred for the randomized and matched video sequences separately (right y-axis). B) shows the change between the rating (t) after a certain video and the previous rating (t - 1) averaged over participants and all trials, including videos from the same video category. Subplot B has the same emotion categories as subplot A (left label of A). The bar represents the mean of the change in ratings, the black line shows the standard deviation, and the * indicates a significant difference from zero. The black frame shows the dominant emotion for the specific emotion category (the emotion intended to be elicited by watching videos from that category). The unframed bars show that videos from a specific emotion category also affected non-dominant emotions. C) shows the autocorrelation coefficient averaged over participants for five lags for each emotion. The line indicates the mean and the shaded area standard deviation over participants. D) shows the mean and standard deviation of the cross-correlation coefficients between emotion time-series averaged over participants. Significance *≤ 0.5, **≤ 0.1, ***≤ 0.001, ****≤ 0.0001

Dynamic effects can also be observed in the cross-correlation between emotion time-series showing that emotions interact with each other (|r| ∈ [0.28 - 0.52], all p < 0.001; cf. Fig. 3D and Supplementary Material Table I.4).

Finally, the test-retest reliability of emotions elicited by repeated video clips varied accordingly to emotion type. Specifically, it was observed to be good for Amusement/Joy (video sequence 1: ICC(2, 1) = 0.79, CI = [0.69,0.86], p < 0.001, video sequence 2: ICC(2, 1) = 0.8, CI = [0.69, 0.87], p < 0.001) and moderate for Disgust/Horror (video sequence 1: ICC(2, 1) = 0.59, CI = [0.43, 0.72], p < 0.001, video sequence 2: ICC(2, 1) = 0.59, CI = [0.43, 0.73], p < 0.001).

In summary, the experimental setup resulted in reliable elicitation of complex, high-dimensional emotional states, allowing for the joint characterization of emotional input sensitivity and emotion dynamics.

3.3 Establishing a Dynamical Model

We next examined the dynamical properties of the emotion ratings, and the interaction with the emotion inputs using a Kalman Filter. First, we compared different models based on the group-level BIC to evaluate which dynamical components are required to capture the data over the whole experiment, i.e. the concatenated time-series before and after the intervention (Fig. 4A). We found that the most parsimonious model, carrying 99% of the cumulative model weight, included a dynamics matrix A, a full input weight matrix C and diagonal noise covariances Σ and Γ. Importantly, data generated from this model accurately captured the observed data sequences as shown in the blue-shaded part in figure 4C. Hence both external inputs and intrinsic dynamics are required to explain the self-reported emotion ratings. The recoverability of the matrices A and C in the most parsimonious model is shown in Supplementary Materials F Parameter Recovery.

Model Evaluation

A) shows the differences in Bayesian Information Criterion (BIC) scores for all models tested compared to the most parsimonious model (blue star). All models were separately fitted to individuals’ emotion rating time-series over the whole experiment. The left y-axis shows the number of free parameters for each model. The most parsimonious model included a dynamics matrix, input weights and diagonal noise covariances for observation and process noise. B) shows differences in BIC scores of models, allowing for parameters to change after the intervention. While in the control group, a model in which all parameters stayed the same best explained the data (grey star), in the distancing group dynamics matrix, input weights and noise covariances differed in the most parsimonious model (purple star). C) shows empirical (blue) and simulated data (orange) from two randomly selected participants. One of those was allocated to the control intervention and one to the distancing intervention. We simulated data from a linear state space model using the parameter estimates derived from fitting the Kalman Filter to individuals’ emotion rating time-series. The vertical grey dashed line indicates that a video was shown at that time-point which stems from the category aiming to elicit that specific emotion. Blue shading indicates the period before, and green shading after the intervention. Data was simulated based on different parameter sets before and after the intervention.

3.4 Effects of Distancing on Emotional Responses

We first examined the effect of the intervention on the multivariate reported emotions averaged over trials (Fig. 5A). The mean ratings averaged over the video block after the intervention were significantly different between the distancing and the control group (T2 = 24.48, F = 4.71, p < 0.001). In contrast, before the intervention the mean ratings did not show a significant difference between groups (T2 = 6.53, F = 1.26, p = 0.29). The within-subject change in average ratings (after minus before intervention) differed significantly between groups(T2 = 52.66, F = 10.14, p < 0.001). Additionally, we observed a significant interaction between time and intervention group (F (5, 208) = 2.6, p = 0.026).

At the group-level, all emotion ratings were significantly reduced after the distancing intervention (U ∈ [688,910] for all comparisons, all p ≤ 0.001), except for ratings of calmness which were increased (U = 1824, p = 0.02; though this does not survive Bonferroni correction p ≤ 0.055 ≤ 0.01). Furthermore, while this pattern was observable in the distancing group (W ∈ [68, 308] for all comparisons, all p < 0.001, before vs after), no changes were detectable in the control group (W ∈ [401,561] for all comparisons, all p ≥ 0.01, except for amused W = 273.0, p < 0.001).

Distancing also affected the temporal variability in emotion ratings (group comparison after intervention: T2 = 36.53, F = 7.03, p < 0.001). The variances of all emotions were reduced (U ∈ [576, 983] for all comparisons, all p ≤ 0.004) in the distancing group compared to the control group and they significantly differed before and after the intervention within the distancing group (W ∈ [133, 246] for all comparisons, all p < 0.001). However, the interaction effect in a MANOVA was not significant (F (5, 208) = 1.72, p = 0.13).

Overall, the distancing intervention strongly affected the mean emotions subjects reported and the variability of emotions (cf. full table in Supplementary Materials Table I.5).

To examine whether the observed group differences are attributable to a demand effect, we analyzed the time taken by participants to rate their emotions on the sliders. We reasoned that if a demand effect were present, participants would respond more quickly as they would already have decided in advance of the question how to respond. Using a mixed ANOVA with time (before vs after the intervention) and intervention groups as factors, we found evidence of a significant interaction effect between time and group (F (1, 106) = 6.46, p = 0.01). However, interestingly, the control group demonstrated a decrease in reaction time (before: M = 12029, SD = 2813 (ms); after: M = 10964, SD = 2713 (ms); t = 5.87, p < 0.001), while the distancing group did not show a significant difference in reaction time before and after the intervention (before: M = 11876, SD = 2796 (ms); after: M = 11518, SD = 2896 (ms), t = 1.73, p = 0.09).

3.5 Effects of Distancing on Dynamics of Emotions

Next, we examined whether the intrinsic dynamics or the input weights or both were altered by the distancing intervention (Fig. 4C). In principle, the observed effect of distancing on the mean and variance of emotion ratings can be explained either by changing the internal dynamics or by changing the input weights alone (cf. Supplementary Material Figure G.5). We hence applied model comparison to examine whether either a change in the dynamics, or in the input weights, or in both was required to capture the data. In the control group, the most parsimonious model (99% of model weight) was the one where the dynamics before and after the intervention stayed the same. That is, there was no evidence for a change in either input weights or dynamics. By contrast, in the distancing group, a model where noise, dynamics, and input weights changed with the intervention provided the most parsimonious account of the data (99% of model weight). Again, data generated from this model was able to capture the observed data sequences accurately (cf. Fig. 4D). Overall, this suggests that the distancing intervention changed both the dynamics and the influence of the videos.

We then examined the specific dynamical features that changed in response to the distancing intervention. Four participants had to be excluded as they were outliers in dynamical characteristics (cf. Supplementary Materials E Exclusion). The following analyses were hence based on 104 participants (54 randomized to the distancing intervention).

Distancing altered the input weights (C matrix) shown in a group comparison of the matrix norm (after intervention: distancing group M = 256, SD = 218; control group M = 400, SD = 222; U = 1897, p < 0.001) and in Figure 5B&C.

Furthermore, a linear dynamical system can be decomposed into eigenmodes - parallel, independent dynamical systems - using an eigendecomposition of the dynamics matrix A (cf. 2.3.2 Stability). Distancing altered the composition of the emotional eigenmodes (T2 = 14.81, F = 2.85, p = 0.019; Fig. 5D). This component was also more stable (decayed more slowly) in the distancing group (distancing group M =0.69, SD = 0.21; control group M = 0.52, SD = 0.22; U = 762, p < 0.001; Fig. 5D) after the intervention. Hence, when controlling for the emotional input, the distancing intervention had specific effects on how different emotional states persisted and interacted.

3.6 Effects of Distancing on Controllability of Emotions

The dynamics matrix A and the input weight matrix C jointly determine the extent to which the emotional state can be controlled by external inputs. This can be formally assessed through a measure called controllability, which we turn to next (cf. 2.3.3 Controllability). Controllability formalizes how strong the inputs to the system have to be to move the dynamical system around, i.e. how ‘reactive’ the dynamical system is to inputs.

The intervention altered controllability overall (group comparison of matrix norm after intervention: distancing group M = 335, SD = 284; control group M = 509, SD = 292; U = 1851, p = 0.001).

Controllability can, however, vary, with the system being more controllable in certain directions than others. This can be examined by studying the eigenspace of the controllability Gramian C, which is based on a combination of dynamics and input matrices, A and C. The eigenspace corresponding to the dominant eigenvalue of the controllability matrix, describing the most controllable direction, differed between the intervention groups after (T2 = 24.67, F = 4.74, p < 0.001), but not before (T2 = 3.69, F = 0.71, p = 0.618), the intervention. A posthoc MANOVA revealed a significant interaction between time and intervention group on the most controllable direction (F (5, 200) = 2.5, p = 0.03). Distancing altered which combination of emotions was most controllable (Fig. 5F), with combinations involving more calm (U = 874, p = 0.002) and less sad emotions (U = 1926, p < 0.001) being most controllable. This direction was, overall, less controllable (lower singular value; U = 1844, p = 0.001; Fig. 5E) in the distancing group.

Intervention Effects

A) shows the average ratings separated in before and after the intervention and for both intervention groups.

Purple shades refer to the distancing group and grey shades to the control group. The boxes show the quartiles of the ratings, while the whiskers extend to show the rest of the distribution. The horizontal bars in the boxes indicate the median, and the green triangles the mean across participants. B) shows mean and standard deviation for the elements of the input weight matrix (C matrix) averaged across participants before the intervention occurred. C) shows the effect of distancing on the input weight matrix (C matrix), which is computed as the mean change (after minus before intervention) divided by its standard deviation for each matrix element only in participants allocated to the distancing intervention. D) The boxplot presents the quartiles of the dominant eigenvalues of the dynamics matrix, while the whiskers display the rest of the distribution. E) The boxplot illustrates the quartiles of the dominant singular values of the controllability matrix, with the whiskers showing the rest of the distribution. F) The polar plots show the unit vector direction of the dominant singular value separated between participants in the distancing (purple shades) and the control (grey shades) group before (left) and after (right) intervention. Significance *≤ 0.05, **≤ 0.01, ***≤ 0.001, ****≤ 0.0001. We report the significance after Bonferroni correction for testing five different emotions.

3.7 Psychopathological correlates

The control of emotional states is thought to be altered in mental illness. Hence we examined controllability in relation to the three assessed psychological questionnaires, focusing on the self-reported emotion regulation difficulties. All questionnaires showed a good internal consistency (Ca ∈ [0.89,0.91]). The total scores of the three questionnaires were highly correlated (r ∈ [0.69,0.75], p < 0.001) and distributions of scores were skewed towards lower values (cf. Supplementary Material Figure H.6).

Psychological well-being was significantly related to emotion ratings at baseline (t0), emotion ratings averaged over the first video block (t1) and the second video block (t2). Ratings of disgust, sadness and anxiousness were positively associated with PHQ-9, GAD-7, and DERS-18 total scores (β ∈ [0.18,0.56] for all comparisons, all p ≤ 0.03; except for GAD-7 and disgust at t0 and t2 β < 0.12, p > 0.2). Ratings of calmness decreased with increasing symptoms (β ∈ [-0.18, -0.4] for all comparisons, all p ≤ 0.05). There was no relationship between amusement and symptoms. However, critically none of the symptom scores interacted with group allocation in predicting change in emotion ratings (all p > 0.05; cf. Supplementary Material Table I.7).

Psychopathological symptoms were related to aspects of the dynamics of emotions. Difficulties in distancing as measured by the DERS-18 were related to the most controllable direction in the emotion space. In those participants with high DERS-18 scores, the most controllable direction pointed towards disgust (β = 0.26, p = 0.006), and away from amusement (β = -0.26, p = 0.005) and calmness (β = -0.24, p = 0.011; though this did not survive Bonferroni correction). In other words, self-reported emotion regulation difficulties were related to needing less effort to drive disgust and more effort to steer the emotional experience towards calm and amused. The effect of DERS-18 total score on disgust was even more prominent after the intervention (β = 0.35, p < 0.001).

There was no evidence that the symptoms moderated the effect of distancing (all p > 0.05; cf. Supplementary Material Table I.8).

4 Discussion

Psychotherapeutic approaches effectively treat a variety of psychiatric conditions. However, the mechanisms underlying these psychotherapeutic effects remain unclear (Kazdin, 2009). By isolating a specific intervention component, such as emotion regulation, we aim to gain a deeper understanding of the processes involved in this specific treatment effect. Emotion regulation techniques are core components of psychotherapy (McRae and Gross, 2020; Powers and LaBar, 2019; Gross, 2015). To study emotion regulation strategies, it is essential to investigate emotions and consider that they fluctuate in intensity and frequency over time, potentially giving rise to a complex dynamical system. We believe that by examining the emotion dynamics of a particular individual, we can better understand how emotions change and how they respond to different interventions.

Here, we studied the impact of an emotion regulation technique called distancing on the dynamics of emotions over time. We used rich and powerful video stimuli to evoke and alter a complex emotional state in participants. This approach allowed us to achieve three main objectives: first, we established how different emotions influence each other over time in a dynamic system; second, we assessed the relative impact of the video clips on these emotions; and finally, we disentangled the effect of distancing on the stability and controllability of these emotional states.

Importance of external inputs in a dynamical system

Our findings revealed that video clips evoked complex patterns in self-reported emotional experiences, including interrelated emotional trajectories. This highlights the significance of considering multidimensional emotional experiences, their interactions, and temporal dependencies, rather than focusing on a single emotion in isolation (Kendler, 2016; Hitchcock et al., 2022; Durstewitz et al., 2020; Kuppens and Verduyn, 2017; Lange et al., 2022). Computational modeling results suggest that external inputs (video clips) were important in explaining a higher dimensional emotional state. In other words, models that included these inputs provided a better explanation of the self-reported emotion responses. This is in line with recent experimental research demonstrating how unexpected and personally-relevant events are associated with measurable changes in the time course of individuals’ emotional responses (Villano et al., 2020; Rutledge et al., 2014; Eldar et al., 2016; Asutay et al., 2022; Lapate and Heller, 2020). Omitting inputs when studying the properties of a dynamical system can lead to inaccurate conclusions as the inputs can mask or alter the apparent dynamics of the system. For instance, Vanhasbroeck et al. (2022) has observed that nonlinearity observed in affective time series in some individuals was the result of external inputs rather than underlying nonlinearity in affect. Hence, to ensure an accurate understanding of the affective system, it is crucial to consider the relationship between affect dynamics and the immediate environment.

Distancing effect on emotional controllability

First, the distancing intervention reliably reduced ratings and variability of both positive and negative emotions but increased ratings of a specific variable, namely feeling calm.

Second, the modeling approach enabled a formal, quantitative assessment of the impact of the distancing intervention on these components. Model comparison showed that the combination of dynamic and controllability signatures changed due to the distancing intervention. Specifically, a brief distancing intervention stabilized intrinsic emotional dynamics and reduced the impact of external stimuli. This finding is significant because static mean differences could be attributed to changes in either dynamics or input sensitivity alone. Interestingly, the intervention affected emotions differently: the controllability of calmness decreased, while sadness increased relative to other emotions. This suggests that some emotions may be more responsive to regulation through distancing.

Emotional controllability and psychopathology

Psychopathology has been associated with deficits in cognitive control (Grahek et al., 2018; Snyder, 2013). The belief that emotions can be controlled to some extent was linked to a decrease in symptoms of anxiety and depression. This connection can be attributed, at least in part, to the fact that individuals with such beliefs tend to employ adaptive strategies for regulating their emotions more frequently (Somerville et al., 2022).

Distancing involves several cognitive control processes, such as taking a step back from a situation, observing it objectively, and cognitively re-framing the experience to focus on positive aspects rather than negative feelings (McRae and Gross, 2020; Ochsner et al., 2004; Dorfel et al., 2014; Staudinger et al., 2009). These processes may include inhibiting pre-potent evaluations, shifting attention away from external stimuli towards the self, and maintaining an intention to detach (McRae and Gross, 2020; Ochsner et al., 2004; Dorfel et al., 2014; Staudinger et al., 2009). Our study demonstrated that the distancing intervention effectively reduced emotional variability and increased feelings of calmness, indicating enhanced emotional control. Furthermore, our modeling showed that the distancing intervention stabilized intrinsic emotional dynamics and reduced the impact of external stimuli. This suggests that distancing can improve emotional stability by enhancing cognitive control. This is consistent with evidence that individuals with high self-control are more successful in regulating emotions (Paschke et al., 2016) and that emotional stability is linked to self-control (Daly et al., 2014; Tangney et al., 2004; Oaten and Cheng, 2006).

An interesting consideration is that cognitive biases may influence emotion regulation ability, thereby contributing to sustained negative affect and reduced levels of positive affect (Joormann and Tanovic, 2015; McRae et al., 2012). For instance, individuals with depression often have difficulty inhibiting negative thoughts or shifting their attention away from negative stimuli. This can lead to a cycle of negative thinking that worsen symptoms (Beck and Haigh, 2014). Our study found that difficulties in emotion regulation were specifically linked to the controllability of emotions such as disgust, amusement, and calmness, indicating that these emotions may be more intrinsically tied to individuals’ perceived ability to regulate their emotions (Gross, 2015; Mitchell, 2021). Notably, emotions like amusement and disgust are strongly influenced by external events, suggesting that the emotional impact of external stimuli plays a critical role in the complexity of emotion regulation difficulties in psychopathology. However, contrary to our expectation, we did not find a correlation between symptoms of psychopathology and the efficacy of the distancing intervention.

Future directions

Our investigation confirms that emotional dynamics and their controllability are complex and context-dependent. Future research needs to be examined whether the findings discovered in this study can be applied beyond the realm of video-induced emotions. The key challenge here will be the appropriate, but necessary, characterisation of inputs. Unless inputs are well-known, the true internal dynamics are not identifiable. However, in more naturalistic environments this appears very difficult (Malamud and Huys, 2024).

4.1 Limitations

This study comes with several limitations. First, it is challenging to identify the degree of influence of possible demand effects. We attempted to avoid this influence by providing participants with the sense that distancing does not work for everyone, and we were interested in how it works for them; however, this might not have been enough. On the other hand, the specifics of the effects of the intervention are probably not predicted from simple demand effects.

Second, employing a standard Kalman Filter approach to analyze the time-series of emotion ratings and video inputs, was based on several considerations. i) There are ample tools for the analysis and characterization of Kalman Filters readily available. ii) The Kalman Filter allows for observations to be noisy. This contrasts with typical analyses of EMA and similar timeseries emotion ratings data, which do not allow for noise in the observations. This is important because the noise in ratings influences estimates of parameters as the rating error ‘persists’ in the modelled future. Nevertheless, one drawback of the Kalman Filter approach we employed was that Gaussian observation noise was assumed, and this is likely to impact the details of the results. However, reconstructed time-series were qualitatively close to the real time-series, and hence we do not believe that the fundamental conclusions here are likely to be affected by this.

Finally, it is possible that eye gaze redirection away from emotionally charged regions in the videos could have acted as an intervening factor in emotion downregulation. Future research could tackle that problem by tracking eye movement.

4.2 Conclusion

In conclusion, we used a novel methodological approach to characterize a key aspect of emotional states, namely how they can be controlled. We found that a brief distancing intervention can effectively regulate emotional experiences, reduce emotional variability and enhance the feeling of calmness. Moreover, distancing leads to qualitatively different changes in the dynamical structure of emotional states. First, it increases the intrinsic emotional stability, and second it reduces the impact of external inputs. Together, these alter the extent to which emotional states are externally controlled. Further computationally detailed characterization of emotional state dynamics and psychotherapeutic interventions may be useful on the path towards understanding differential and specific effects of different psychotherapeutic interventions.

eLife Assessment

This important manuscript proposes a dual behavioral/computational approach to assess emotional regulation in humans. The authors present solid evidence for the idea that emotional distancing (as routinely used in clinical interventions for e.g. mood and anxiety disorders) enhances emotional control.

Reviewer #1 (Public review):

Summary:

Using sequences of short videos to elicit emotional changes in participants, Malamud & Huys demonstrate how a brief, controlled emotion regulation intervention (distancing) can effectively alter subsequent emotion ratings. The authors employ latent state-space models to capture the trajectories of emotion ratings, leveraging tools from control theory to quantify the intervention's impact on emotion dynamics.

Strengths:

The experiment is well-designed and tailored to the computational modeling approach advanced in the paper. It also relies on a selection of stimuli that were previously validated. Within the constraints of a controlled experiment, the intervention successfully implements a relatively common tool of psychotherapeutic treatment, ensuring its clinical relevance.

The computational modeling is grounded in the well-established framework of dynamical systems and control theory. This foundation offers a conceptually clear formalization along with powerful quantification tools that go beyond previous more data-driven approaches.

Overall, the study presents a coherent approach that bridges concepts from clinical psychology and computational theories, providing a timely stepping stone toward advancing quantified, evidence-based psychological interventions targetting emotion control.

Weaknesses:

A primary limitation of this study, acknowledged by the authors, is its reliance on self-reports of participants' emotional states. Although considerable effort was made to minimize expectation effects, further research is needed to confirm that the observed behavioral changes reflect genuine alterations in emotional states. Additionally, the generalizability of the findings to long-term remediation strategies remains an open question.

Second, the statistical analysis, particularly the computational approach, sometimes lacks sufficient detail and refinement. While I will not elaborate on specific points here, one notable issue is the interpretation of the intrinsic matrix (A). The model-free analysis reveals correlations between emotions at a given time or within an emotional state across time points. However, it does not provide evidence to support lagged interactions across states that would justify non-diagonal elements in A. The other result concerning the dynamics matrix only highlights a trend in the dominant eigenvalue, which is difficult to interpret in isolation. The absence of a statistically significant group x intervention interaction furthermore makes this finding a little compelling. This weakens the study's conclusions about the importance of intrinsic dynamics, as claimed in the title.

Finally, to avoid potential misunderstandings of their work, the authors should be more careful about their use of terms pertaining to the control theory and take the time to properly define them. For example, the "controllability" of emotional states can either denote that those states are more changeable (control theory definition), or, conversely, more tightly regulated (common interpretation, as used in the abstract). This is true for numerous terms (stability, sensitivity, Gramian, etc.) for which no clear definition nor references are provided. Readers unfamiliar with the framework of control theory will likely be at a loss without more guidance.

Reviewer #2 (Public review):

Summary:

In this well-conceived and timely study, the authors assess the controllability of emotions in a quantitative way using the framework of control theory. They use a controlled distancing intervention halfway through an emotion rating task where emotion-inducing short videos from a validated database are shown and find that the intervention enables a better controllability of externally induced emotions in the experimental group.

Strengths:

It is a highly original idea to address the external controllability of emotions using the formal framework of control theory. It is also a very propitious approach to take what could be called a 'micro-therapeutic' perspective which looks at the immediate effect of an intervention instead of the 'macro-therapeutic' mid- or long-term effect of a whole course of therapy.

Weaknesses:

Acquiring data online inevitably gives rise to selection and self-selection effects. This needs to be acknowledged clearly. Exacerbating this, participant remuneration seems low at an amount below the minimum or living wage in Western countries (do the authors know where their participants came from?).

Another concern is that the intervention does not simply take place before the second block begins but is ongoing during the whole of the second block in that it is integrated into the phrasing of the task on each trial. It is therefore somewhat misleading to speak of a period 'after the intervention', and it would have been interesting to assess the effect of this by including a third group where the phrasing does not change, but the floating leaves intervention takes place.

As mentioned in the Limitations section, observation noise was assumed and not estimated. While this is understandable in this case, the effect of this assumption could have been assessed by simulation with varying levels of observation (and process) noise.

Relatedly, the reliance on formal model comparison is unfortunate since the outcome of such comparisons is easily influenced by slight changes to assumptions such as noise levels. An alternative approach would have been to develop a favoured model based on its suitability to address the research question and its ability, established by simulation, to distill relevant changes of behaviour into reliable parameter estimates.

The statistical analyses clearly show the limitations of classical statistical testing with highly complex models of the kind the authors (commendably) use. Hunting for statistically significant interactions in a multivariate repeated-measures design relying on inputs from time series-derived point estimates is a difficult proposition. While the authors make the best of the bad situation they create by using null-hypothesis significance testing, a more promising approach would have been to estimate parameters using a sampler like Stan or PyMC and then draw conclusions based on posterior predictive simulations.

Reviewer #3 (Public review):

Summary:

The manuscript takes a dynamical systems perspective on emotion regulation, meaning that rather than a simplistic model conceptualising regulation as applying to a single emotion (e.g. regulation of sadness), emotion regulation could cause a shift in the dynamics of a whole system of emotions (which are linked mathematically to one another). This builds on the idea that there are 'attractor states' of emotions between which people transition, governed by both the system's intrinsic characteristics (e.g. temporal autocorrelation of a particular emotion/person) and external driving forces (having a stressful week). Conceptually this is a very useful advance because it is very unlikely that emotions are elicited (or reduced) singly, without affecting other emotions. This paper is a timely implementation of these ideas in the context of psychotherapeutic intervention, distancing, which participants were trained (randomised) to perform while watching emotion-inducing videos.

The authors' main conclusion is that distancing both stabilises specific emotional patterns and reduces the impact of external video clips. I would consider these results strong and believable, and to have the potential to impact models of emotion regulation as well as the field's broader views on the mechanisms of psychological therapies.

Strengths:

This paper has very many strengths: I would especially note the authors' very-well-matched active control condition and the robustness of their model comparison approach. One feature of the authors' approach is that they explicitly add noise - not what you typically see in an emotion time-series analysis - which allows participants to make errors in their own subjective ratings (a reasonable thing to assume); this noise can then be smoothed during filtering. In their model comparison approach, they explicitly test whether a true dynamical system explains emotion change/emotion regulation effect on emotions - demonstrating that both intrinsic dynamics and external inputs were needed to explain subjective emotion. Powerfully, they also used this approach to test the differential effects of the treatment groups (see below).

The main result seems quite robust statistically. Verifying the effects of the distancing intervention on emotion, the authors found an interaction between time (pre- to post-intervention) and intervention group (distancing vs. relaxation) suggesting that distancing (but not relaxation) reduced ratings of almost all emotions. Participants allocated to the distancing intervention also showed decreased variability of emotion ratings compared to those in the relaxation intervention (though note this interaction was not significant).

Using a model comparison approach, the authors then demonstrated that whilst the control group was best explained by a model that did not change its dynamics of emotions, the active intervention (distancing) group was best explained by a model that captured both changing emotion dynamics and a changing input weights (influence of the videos) - results confirmed in follow-up analyses. This is convincing evidence that emotion regulation strategies may specifically affect the dynamics of emotions - both their relationships to one another and their susceptibility to changes evoked by external influences.

The authors also perform analyses that suggest their result is not attributable to a demand effect (finding that participants were quicker during the control intervention, which one would expect if they had already decided how to respond in advance of the emotion question). I personally also think a demand effect is unlikely given the robustness of their control intervention (which participants would be just as likely to interpret as mental health-enhancing training as distancing), and I am convinced by the notion that demand effects would be unlikely to elicit their more specific effects on the dynamic quality of emotions.

Weaknesses:

An interesting but perhaps at present slightly confusing aspect of their described results relates to the 'controllability' of emotions, which they define as their susceptibility to external inputs. Readers should note this definition is (as I understand it) quite distinct from, and sometimes even orthogonal to, concepts of emotional control in the emotion literature, which refer to intentional control of emotions (by emotion regulation strategies such as distancing). The authors also use this second meaning in the discussion. Because of the centrality of control/controllability (in both meanings) to this paper, at present it is key for readers to bear these dual meanings in mind for juxtaposed results that distancing "reduces controllability" while causing "enhanced emotional control".

As above the authors use an active control - a relaxation intervention - which is extremely closely matched with their active intervention (and a major strength). However, there was an additional difference between the groups (as I currently understand it): "in the group allocated to the distancing intervention, the phrasing of the question about their feelings in the second video block reminded participants about the intervention, stating: "You observed your emotions and let them pass like the leaves floating by on the stream." I do wonder if the effects of distancing also have been partially driven by some degree of reappraisal (considered a separate emotion regulation strategy) since this reminder might have evoked retrospective changes in ratings.

Not necessarily a weakness, but an unanswered question is exactly how distancing is producing these effects. As the authors point out, there is a possibility that eye-movement avoidance of the more emotionally salient aspects of scenes could be changing participants' exposure to the emotions somewhat. Not discussed by the authors, but possibly relevant, is the literature on differences between emotion types on oculomotor avoidance, which could have contributed to differential effects on different emotions.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public reviews):

Weaknesses:

Reliance on self-reports

A primary limitation of this study, acknowledged by the authors, is its reliance on self-reports of participants’ emotional states. Although considerable effort was made to minimize expectation effects, further research is needed to confirm that the observed behavioral changes reflect genuine alterations in emotional states. Additionally, the generalizability of the findings to long-term remediation strategies remains an open question.

We agree with this characterisation and have strengthened the corresponding acknowledgment in the Discussion. We would also note that, while self-report measures are inherently subjective, the regularities governing subjective emotional experience are of primary scientific interest in their own right, and no consensus definition of a ”genuine” emotional state that supersedes self-report currently exists. We have added a sentence to the Discussion making this point explicit: ”While emotional self-reports are inherently subjective, we note that the regularities governing subjective emotional experience are of primary scientific interest in their own right, and no consensus definition of a ‘genuine’ emotional state that supersedes self-report currently exists.”

Additionally, we agree that what we have described is limited to a short-term intervention and change. Whether these changes bear on longer-term changes remains to be assessed. Furthermore, the mechanisms or processes that would support such a maintenance are of substantial interest, and will be the focus of future work.

Statistical analysis and interpretation of the dynamics matrix

Second, the statistical analysis, particularly the computational approach, sometimes lacks sufficient detail and refinement. While I will not elaborate on specific points here, one notable issue is the interpretation of the intrinsic matrix (A). The model-free analysis reveals correlations between emotions at a given time or within an emotional state across time points. However, it does not provide evidence to support lagged interactions across states that would justify non-diagonal elements in A. The other result concerning the dynamics matrix only highlights a trend in the dominant eigenvalue, which is difficult to interpret in isolation. The absence of a statistically significant group x intervention interaction furthermore makes this finding a little compelling. This weakens the study’s conclusions about the importance of intrinsic dynamics, as claimed in the title.

We thank the reviewer for raising this important methodological point. We address it in three parts.

(i) Justification for the full dynamics matrix. In response to this comment, we have added a diagonal-constrained variant of A to the model comparison. The full A model was selected by BIC over the diagonal-constrained version, meaning that cross-emotion lagged interactions contributed to model fit over and above what could be explained by individual emotion autocorrelation alone, thereby justifying the inclusion of off-diagonal elements. We have updated the model comparison results section and figure caption accordingly.

(ii) Interpretation of the dominant eigenvalue. We agree that the dominant eigenvalue alone is difficult to interpret. Our primary claim regarding intrinsic dynamics rests on model comparison (which identified a change in A as necessary to explain post-intervention data in the distancing group) together with the change in the direction of the dominant eigenvector (tested via Hotelling T2, p = 0.019). The eigenvalue magnitude is reported as a complementary, interpretable summary of the stability shift.

(iii) General analysis of the dynamics matrix. We have added a visualization of the full dynamics matrix A to the Supplementary Materials, reporting per-emotion diagonal elements (persistence) and their variability across participants. This shows that the emotion calm exhibited the highest persistence, consistent with our hypothesis, whereas sadness showed lower-than-expected stability, and other emotions exhibited low mean persistence with substantial individual variability. We have added the following sentence to the Results: ”A general analysis of the dynamics matrix A revealed that while calm exhibited the highest persistence, consistent with our hypothesis (0.59 ± 0.85), sadness did not show the expected stability (0.09 ± 0.58), and other emotions demonstrated low persistence (0.13 to 0.24) alongside substantial individual variability.”

Terminology (controllability, stability, sensitivity, Gramian)

Finally, to avoid potential misunderstandings of their work, the authors should be more careful about their use of terms pertaining to the control theory and take the time to properly define them. For example, the ”controllability” of emotional states can either denote that those states are more changeable (control theory definition), or, conversely, more tightly regulated (common interpretation, as used in the abstract). This is true for numerous terms (stability, sensitivity, Gramian, etc.) for which no clear definition nor references are provided. Readers unfamiliar with the framework of control theory will likely be at a loss without more guidance.

This is an excellent and important point. We have made the following changes throughout the manuscript.

(i) Terminology table. We have added a new Table 1 in the Methods (Conceptual Definitions) providing a side-by-side mapping of the formal control-theory definition and the psychological interpretation for the three key terms: Controllability, Stability, and Sensitivity.

(ii) Abstract. The abstract previously used ”controllability” in a way that could be read in the psychological sense. We have replaced the sentence referencing the Controllability Gramian with a clarified version describing the measure as ”continuous measures derived from the controllability matrix, capturing relative differences in how emotional states can be driven by external inputs.”

(iii) Introduction. We have added a paragraph explicitly distinguishing the control-theoretic definition of controllability from the psychological concept of emotion control or emotion regulation, noting that high controllability in the formal sense does not imply tight regulation or reduced variability. Definitions of sensitivity and stability are also provided.

(iv) Discussion. Occurrences of ”controllability” that could be ambiguous are annotated with brief clarifications of which sense is intended.

(v) Gramian. We now use the ”controllability matrix” (C) throughout, and its SVD-based analysis is described using ”left singular vectors” rather than ”eigenvectors of the Gramian”.

Reviewer #2 (Public reviews):

Online recruitment, selection effects, and remuneration

Acquiring data online inevitably gives rise to selection and self-selection effects. This needs to be acknowledged clearly. Exacerbating this, participant remuneration seems low at an amount below the minimum or living wage in Western countries (do the authors know where their participants came from?).

We thank the reviewer for raising this. All participants were recruited from the UK via Prolific Academic and reimbursed at £7.50/h. Remuneration rates were comparable to other experimental settings, in keeping with other online studies, UK living wage recommendations, and ultimately determined according to institutional ethical guidance. We acknowledge that online recruitment via Prolific may introduce self-selection effects and that the sample may not be representative of the general population. We have added a corresponding statement to the Limitations section: ”Online recruitment via Prolific Academic may introduce self-selection effects, and the sample may not be representative of the general population.”

Intervention ongoing during the second block

Another concern is that the intervention does not simply take place before the second block begins but is ongoing during the whole of the second block in that it is integrated into the phrasing of the task on each trial. It is therefore somewhat misleading to speak of a period ’after the intervention’, and it would have been interesting to assess the effect of this by including a third group where the phrasing does not change, but the floating leaves intervention takes place.

This is a valid and important observation. In the distancing group, the trial-by-trial question phrasing during the second block included a reminder, meaning the intervention was reinforced on every trial. We have acknowledged this as a procedural difference and potential confound in the Limitations section, noting that this reminder may have encouraged a form of retrospective reappraisal rather than in-the-moment distancing. The design choice was intentional (the reminder was included to encourage continued application of the strategy, mirroring how such techniques are deployed in practice) but we agree it is an imperfect feature of the current design and have discussed it openly. We have added to Limitations: ”Relatedly, a procedural difference existed in that only the distancing group received a strategy reminder during the second video block. While intended to facilitate real-time regulation, this prompt may have inadvertently encouraged retrospective reappraisal to align with the distancing narrative.”

Observation noise

As mentioned in the Limitations section, observation noise was assumed and not estimated. While this is understandable in this case, the effect of this assumption could have been assessed by simulation with varying levels of observation (and process) noise.

We would like to clarify that both observation noise (Γ) and process noise (Σ) were in fact estimated from the data, constrained to be diagonal. We have extended the parameter recovery analysis (Supplementary Section: Parameter Recovery) in which, for each of 104 subjects and both time periods (N = 208 observations), we generated 100 surrogate trajectories from the fitted model and re-estimated all parameters. Recovery quality is quantified via Pearson correlations between true and recovered parameters (A, C, Σ, and Γ), alignment of dominant eigenvectors and left singular vectors, and bias analysis for dominant eigenvalue and singular value magnitudes. The analysis confirms that A and C matrix parameters and the critical eigenvector-based metrics all exceed our target threshold of r = 0.7. Noise covariances are poorly recovered, consistent with typical Kalman filter behaviour on short time series. We have expanded the Limitations to note that the Gaussian observation noise assumption does not fully capture the bounded nature of 0–100 rating scales, and suggest that future work could use a truncated or censored observation model.

Reliance on formal model comparison

Relatedly, the reliance on formal model comparison is unfortunate since the outcome of such comparisons is easily influenced by slight changes to assumptions such as noise levels. An alternative approach would have been to develop a favoured model based on its suitability to address the research question and its ability, established by simulation, to distill relevant changes of behaviour into reliable parameter estimates.

We appreciate this methodological concern but would argue that formal model comparison is wellsuited to our research question. Our central aim is not simply to fit emotion trajectories, but to determine which components of the dynamical system (intrinsic dynamics (A), input weights (C), or both) are altered by the distancing intervention. This is inherently a model comparison question: without comparing models that do and do not allow each component to change, no principled inference about the mechanism of action is possible. A single favoured model, however well-motivated, would presuppose the answer.

We also note that the reviewers concern, that outcomes are sensitive to underlying assumptions, applies equally to the favoured-model approach: simulations rely on predefined structures and noise specifications that shape parameter recovery, and a misspecified favoured model risks confounding the parameters intended to capture the intervention effect with artefacts of model structure. By contrast, our approach evaluates a principled, nested set of models of increasing complexity, using BIC to penalise unnecessary parameters, which guards against overfitting. The models are intentionally simple (linear, Gaussian, time-invariant) to limit the degrees of freedom available to absorb noise.

Critically, the approach is not model comparison alone. We followed established best-practice procedures for computational modelling, including posterior predictive checks (simulated trajectories closely matched observed data; Fig. 4C), and parameter recovery establishing that the key metrics are reliably recovered (see Supplementary Materials F Parameter Recovery). We have also added a Random-Effects Bayesian Model Selection (RFX-BMS) to characterise individual heterogeneity in model preferences. Together, these provide converging evidence for the validity of our inference. We have clarified this reasoning in the revised manuscript.

Statistical limitations; Bayesian inference

The statistical analyses clearly show the limitations of classical statistical testing with highly complex models of the kind the authors (commendably) use. Hunting for statistically significant interactions in a multivariate repeated-measures design relying on inputs from time series- derived point estimates is a difficult proposition. While the authors make the best of the bad 3 situation they create by using null-hypothesis significance testing, a more promising approach would have been to estimate parameters using a sampler like Stan or PyMC and then draw conclusions based on posterior predictive simulations.

We agree that fully Bayesian parameter estimation via Stan or PyMC would be a valuable methodological advance. Implementing this for 104 subjects across two time periods with a 5-dimensional Kalman filter is, however, a substantial undertaking beyond the scope of the current revision. In the interim, the RFX-BMS analysis added in response to Comment 4 above provides a group-level Bayesian perspective on model uncertainty that partially addresses this concern.

Reviewer #3 (Public reviews):

Dual meanings of controllability

An interesting but perhaps at present slightly confusing aspect of their described results relates to the ’controllability’ of emotions, which they define as their susceptibility to external inputs. Readers should note this definition is (as I understand it) quite distinct from, and sometimes even orthogonal to, concepts of emotional control in the emotion literature, which refer to intentional control of emotions (by emotion regulation strategies such as distancing). The authors also use this second meaning in the discussion. Because of the centrality of control/controllability (in both meanings) to this paper, at present it is key for readers to bear these dual meanings in mind for juxtaposed results that distancing ”reduces controllability” while causing ”enhanced emotional control”

We are grateful for this observation, which echoes Reviewer 1’s concern about terminology. We have addressed this comprehensively; please see our response to Reviewer 1 Comment 3 above.

Strategy reminder and possible reappraisal

As above the authors use an active control – a relaxation intervention – which is extremely closely matched with their active intervention (and a major strength). However, there was an additional difference between the groups (as I currently understand it): ”in the group allocated to the distancing intervention, the phrasing of the question about their feelings in the second video block reminded participants about the intervention, stating: ”You observed your emotions and let them pass like the leaves floating by on the stream.” I do wonder if the effects of distancing also have been partially driven by some degree of reappraisal (considered a separate emotion regulation strategy) since this reminder might have evoked retrospective changes in ratings.

This is a well-founded concern. As noted in our response to Reviewer 2 Comment 2, we have added an explicit acknowledgment of this procedural difference and the potential for retrospective reappraisal to the Limitations section. We note, as the reviewer themselves observe in the Strengths section, that demand effects are unlikely to account for the specific pattern of dynamic changes observed: uniform demand effects would be expected to produce flat reductions across emotions, whereas our findings show emotion-specific changes in eigenmode structure and controllability direction. Nevertheless, a partial contribution of reappraisal cannot be ruled out from the current design.

Mechanism of distancing effects (eye movement and oculomotor avoidance)

Not necessarily a weakness, but an unanswered question is exactly how distancing is producing these effects. As the authors point out, there is a possibility that eye-movement avoidance of the more emotionally salient aspects of scenes could be changing participants’ exposure to the emotions somewhat. Not discussed by the authors, but possibly relevant, is the literature on differences between emotion types on oculomotor avoidance, which could have contributed to differential effects on different emotions.

We thank the reviewer for raising the oculomotor avoidance hypothesis. Research suggests that different emotions elicit distinct patterns of gaze behaviour: disgust is associated with visual avoidance, whereas anxiety and other negative emotions show increased attentional bias following fear conditioning. These emotion-specific oculomotor patterns could have contributed to the differential effects we observe on the input weight matrix C. What would be particularly interesting to examine in future work is whether a distancing intervention induces multiple, emotionally-specific gaze behaviours, or a single undifferentiated avoidance response. We have expanded the Limitations to: ”[...] The literature on emotion-specific oculomotor avoidance suggests that gaze patterns differ across emotion categories, which could contribute to differential effects on the input weight matrix for specific emotions. [...]”

Recommendations for the Authors:

Reviewer #1 (Recommendations for the Authors):

(1) In the procedure description, the authors suggest that some emotions (e.g. disgust) would be more volatile and stimulus driven, while others (eg. sad) would be more stable. Is this hypothesis reflected in the model-based state dynamics, typically in the diagonal elements of A?

Yes, we added a supplementary figure showing the dynamics matrix A, which confirms that calm exhibited the highest persistence, consistent with our hypothesis. Disgust showed lower persistence, also in line with this expectation. However, sadness did not display the expected stability and instead showed relatively low persistence, contrary to our hypothesis.

(2) Could the authors elaborate on the emotional space covered by the chosen ratings? If the axes are positive-negative and slow-fast, why 5 and not 4?

We added a sentence in the Methods clarifying that five emotions were selected based on the specific affective qualities of the video stimuli provided in the validated databaseto and to better capture the high-dimensional, nuanced states elicited by the stimuli rather than to fit a traditional four-axis model.

(3) The whole methods section crucially lacks references. As an example, the whole derivation of the most important metrics (eigenvalues of the Gramian, energy ellipse, etc) leaves the reader completely on its own.

References have been added throughout the Methods, including for the controllability matrix, SVDbased analysis, and eigendecomposition.

(4) Before equation 1, when introducing x and u: a) time appears twice (typo), and b) the 1Tˆ notation is not standard (especially without bold) and unclear until way below when the one-hot encoding is mentioned.

The typo has been corrected.

(5) Why use one hot-encoding rather than the original ratings from the video database? Videos must vary if not in spread (as suggested in Figure B.1) at least in intensity. Ignoring this variance surely diminishes the accuracy of the modelling.

We used one-hot encoding so that the input weight matrix C can directly estimate participantspecific intensity and sensitivity, rather than fixing input magnitudes to database averages. Using database ratings would assume that the emotional intensity of each video clip generalizes perfectly to our sample; any mismatch would be absorbed as error in C, potentially biasing parameter estimates and obscuring individual differences in emotional reactivity, which are central to our analysis.

(6) Could the authors develop the rationale behind the bias in Equation 1?

We added a sentence clarifying that the bias term h captures the steady-state baseline of the emotional system, i.e. the mean rating toward which emotions converge in the absence of external inputs.

(7) While I can understand why the authors included a set of models with a diagonal C matrix, I do not see why they did not do the same with A. While the diagonal elements are necessary to persist emotional states and induce some autocorrelation in the ratings, as observed empirically, the influence of the non-diagonal elements is not justified (and Figure 5G seems to confirm that). This is critical as, in the end, the controllability metrics will highly depend on those non-diagonal elements which remain very obscure throughout the manuscript.

We now included both diagonal and full variants of A in the model comparison; still the full A was selected by BIC.

(8) Concerning the model comparison, I am not sure what the authors mean by using the BIC at the group level. Did they just sum them across participants? This approach is known to be highly susceptible to outliers and cannot be relied upon in general. So-called ”random effect analysis” tends to be regarded as the gold standard and can be easily implemented by taking - 0.5 * BIC as an approximation to the model evidence. Such an approach would also allow to properly test for group differences (cf. Rigoux et al. 2014).

We have added a Random-Effects Bayesian Model Selection analysis as a new Supplementary Section; see Comment 4 of the public review response above.

(9) The sentence ”proportion of the total amount of predictive power provided by the full set of models contained in the model being assessed” does not make any sense to me. Please rephrase.

The sentence has been rephrased: ”Cumulative model weights (wj) normalize raw BIC scores so they can be interpreted as the relative probability that a specific model is the best one among the set being compared:”

(10) ”the largest eigenvalue of the dynamics matrix A identifies the most stable combination of emotions” is only true if the eigenvalues are below 1, which is not granted.

We have added the qualifier that this holds provided the dominant eigenvalue lies within the unit circle (|λ| < 1), indicating a system that converges to a steady state.

(11) Equation 3 does not define the Gramian but the controllability matrix, a confusion that goes through the manuscript. The Gramian is formally defined as W = P(AkBB′A′k). Luckily, for discrete systems, it can be approximated by CC′ and therefore the singular values of C can be used to approximate the eigenvalues of W, which are the usual metrics used to define the energy ellipse and so on. While the results reported in the manuscript are correct (up the the approximation), the general description is wrong or misleading.

We thank the reviewer for this important correction; we now consistently refer to C as the ”controllability matrix” throughout, and its SVD-based analysis is described in terms of left singular vectors rather than eigenvectors of the Gramian.

(12) Figure 2: what is the matrix V? If it’s from the singular value decomposition W = USV, then (if I am not mistaken) the direction of the ellipsoid is defined by U. Again, a reference would help.

We have corrected the figure and caption to refer to ”left singular vectors” throughout. We retain the variable name V rather than adopting the standard SVD convention of U to avoid confusion with the input vector u, which appears throughout the model equations.

<(13) Correction for multiple comparisons is mentioned as a way to correct for the number of conducted tests. However, later on, some post-hoc analyses are reported with the mention that the correction is done across emotions (so p/5), while multiple tests are run for each emotion. This is critical when all the pairs across the cells of an ANOVA are tested and no correction seems to be applied, which is inducing a huge risk of false positives.

Along the same line: the correct way to demonstrate the effect of the intervention is to first do an ANOVA to reveal an interaction between group and time, and then only to do post-hoc tests to pinpoint where the interaction is coming from, and not the other way around as reported in the manuscript. Further, a difference in significance is not equivalent to a significant difference, and showing that a time effect is significant in one group but not in the other does not imply that the intervention differs between groups, only testing the interaction can confirm this.

We have added reporting of the significant group × time interaction effects (F(5,208) = 2.6, p = 0.026 for mean ratings; F(5,200) = 2.5, p = 0.03 for the most controllable direction) and flagged these in the figure captions. In the interest of transparency we have left the structure of the results section intact rather than retrospectively reframing it.

(14) The notation DV = b0 + b1IV*b2G is confusing as a full model (interaction + main effects) should have 3 parameters in addition to the intercept. Also, why use different models for testing the main effect and the interaction?

The regression equation has been corrected to DV = β0 + β1IV + β2G + β3(IV × G) + ϵ, making the interaction term explicit.

(15) Figure 4: the control subject in panel C seems to rate close to 0 in all emotions except for ”calm”. How was this subject fitted? Does model selection (at the subject level) correctly identify a change of dynamics in this case? I don’t see how the behaviour after the intervention could be realistically fitted with a 65-parameters dynamical system. What type of checks were operated to ensure the quality of the fit beyond the recovery analysis (see below)?

The top participant in Figure 4C was from the distancing group and the participant rating close to zero on most emotions except calm shown at the bottom was from the control group. For the distancing participant, model selection correctly identified a change in dynamics and input weights (BIC = 4806) over the same-parameters model (BIC = 4821). For the control participant, model selection similarly favoured a change in dynamics and input weights (BIC = 3723) over the sameparameters model (BIC = 4011), suggesting that the relaxation intervention produced a comparable effect on emotional dynamics to the distancing intervention. Note that these two participants were selected randomly to illustrate the visual quality of model fit (i.e. that simulated trajectories closely resemble the empirical rating curves) and are not intended to be representative examples of group differences.

Regarding the data-to-parameter ratio: the 65 parameters are estimated from 55 observations per emotion per block, giving a more favourable ratio than it might appear.

Beyond the visual trajectory overlays in Figure 4C, we have now added R2 and peak cross-correlation as a quantitative measure of individual fit quality. Mean R2 across all subjects and emotions was 0.6 and mean temporal correlation was r = 0.74−0.80, confirming that both the timing and magnitude of emotional responses were well reproduced. Notably, for the specific control participant shown in Figure 4C, R2 values were 0.74, 0.83, 0.81, 0.83, and 0.83 for disgusted, amused, calm, anxious, and sad respectively, confirming that even for this visually striking participant the model fit was adequate across all five emotion dimensions.

(16) What do the authors mean by ”eigenmodes”? In the following sentence, what does ”This component” refer to?

Eigenmodes is defined in the Stability section as the independently evolving combinations of emotions obtained by projecting the state vector onto the eigenvectors of A, and ”This component” has now an explicit referent.

(17) Figure 5: see above the comment about the necessity for testing the interactions, which should also be reported in the figures.

Interaction effects are now included in the relevant figure caption (Figure 5).

(18) When looking for the relationships between questionnaires and controllability, looking only at the most controllable direction seems rather inefficient due to the multiple comparisons correction. Why not compute the angle (or other measure of similarity) with an ideal ”calm” unit vector?

Along the same line, it’s unlikely that the most controllable direction will smoothly rotate as a function of symptoms. More likely, the winning (highest eigenvalue) direction will switch from one to another, creating some discontinuity in the summary statistic used for the correlation with clinical scores. How could one work around this issue?

This is an interesting suggestion; we have not implemented it in the current revision, but we acknowledge it as a promising analysis for future work.

(19) More generally, it would be interesting to see if there are some regularities in the dynamics across participants. If this is the case, one could construct a canonical emotional dynamics and project all participants on this eigenspace. Emotional trajectories would then differ only in their controllability (eigenvalues) in this common space, making a comparison across participants more straightforward. Could the authors comment on this?

This is a valuable suggestion that we have not pursued in the current revision, as constructing a common eigenspace across participants requires additional methodological choices.

(20) I am a bit puzzled by the hypothesis that questionnaires should mediate the intervention effect. Shouldn’t questionnaires be related to before-intervention controllability only? Similarly, could one use initial controllability to predict the intervention response (irrespective or not of the clinical score)?

Our hypothesis was that participants with greater difficulties in emotion regulation (high DERS-18) would show a smaller intervention effect, as their emotional system might be less amenable to brief distancing. However, this was not the case. We also note that DERS-18 scores were not significantly related to the overall magnitude of pre-intervention controllability (norm of the controllability matrix), but were related to its direction: participants with higher DERS-18 scores showed a most controllable direction pointing toward disgust and away from amusement and calmness, suggesting that trait-level regulation difficulties are linked to the specific emotional configuration of the system rather than its overall controllability.

Regarding using initial controllability to predict the intervention response: we agree this is a mechanistically appealing question, but it is unfortunately not straightforward to address here. The intervention effect would be quantified as the change between pre- and post-intervention controllability, and since pre-intervention controllability is a constituent of that change score, any correlation between the two would be partially circular by construction.

(21) The recovery procedure should be way more detailed. How many surrogates were run, etc.? Which kind of quality checks were used to ensure the recoverability was sufficient at the subject level, especially as parameter recovery seems relatively low for some subjects?

The parameter recovery section now reports that 100 surrogate trajectories were generated per subject (N = 104) per time period (before and after intervention: N = 208 observations total), with per-subject recovery quality reported across simulations together with across-subject variability; see Supplementary Materials F Parameter Recovery.

(22) As the result of the eigen decomposition is the endpoint of the analysis, it would be a nice addition to test the recoverability of those measures (eg. correlation between simulated and inferred eigenvalues).

Recovery of the dominant eigenvalue/vector and dominant singular value/left singular vector is now reported explicitly, including bias analysis and scatter plots of true versus recovered values. Beyond subject-level recovery, we also assessed whether the observed group differences in emotional dynamics and controllability could be reliably recovered at the group level. See Supplementary Materials F Parameter Recovery.

(23) Although this comment comes close to last, this is a major concern of mine. I am not convinced that the recoverability procedure is sufficient to prove that the inference is working. The model assumes that the observation noise is Gaussian, which is clearly not the case in the data. By simulating surrogate time series with normally distributed noise, the authors do not account for any saturating effects that could destroy a large part of the behavioural information necessary for a successful inversion (eg Figure 4.C showing that simulated data contains a lot of negative ratings). A workaround would be to bind the surrogate time series to mimic the saturation caused by the rating scale, and then run the model estimation on those capped time series.

We acknowledge this important limitation: the Gaussian assumption does not capture the bounded 0–100 scale, and we have added this to the Limitations with a suggestion that future work use a truncated or censored observation model.

(24) Supplementary tables with placeholders (v1, v2) that can vary in meaning depending on the line are extremely hard to decipher.

The supplementary tables have been restructured to a hierarchical format.

(25) Table I9 is not referenced in the manuscript.

A reference to this table has been added in the appropriate Results section.

(26) It’s a shame that neither data nor analysis code has been made available.

Fully anonymised data and analysis code are now publicly available on GitHub (https://github.com/huyslab/emotioncon public).

Reviewer #2 (Recommendations for the Authors):

(1) Abstract: By some definitions, controllability is binary, present or absent, according to whether the controllability Gramian is positive definite. Mention that you use a continuous definition, otherwise ’quantified’ leads to confusion.

The abstract now describes the measure as: ”Controllability was assessed using continuous measures derived from the controllability matrix, capturing relative differences in how emotional states can be driven by external inputs.”, making the non-binary usage explicit.

(2) p 5: ’on [not in] the recruiting platform’.

Corrected.

(3) p 8: Clarify notation of xtt = 1T (and same for u). What does this mean?

The notation has been corrected.

(4) p 8: Define h in the paragraph following Eq 1, don’t wait until the next section.

The definition of the bias h (steady-state baseline) has been moved to immediately after Equation 1.

(5) p 9: Give clear references for your methods here. There are different definitions of controllability etc. than the ones you use.

References have been added at each key definition.

(6) p 28: ’20 videos per emotion category were chosen resulting in 50 videos per sequence’ - doesn’t make sense.

This has been clarified: 20 videos per category across 5 categories yield 100 videos in total. These were split into two matched sequences of 50 videos each. Including 2 videos that were repeated twice resulted in 54 videos per sequence and 108 videos in total.

(7) Figure B2: 54 videos are listed, and categorized into five categories. How does this relate to the remark right above?

A clarifying note has been added to the supplementary explaining that each sequence of 54 clips includes repeated videos and is drawn from the pool of 100, with emotion-category sequences matched between blocks.

(8) p 29: What were the process noise Σ and observation noise Γ assumed in the parameter recovery exercise? What were the consequences of that assumption as assessed by simulation?

Both Σ and Γ were estimated from the data constrained to be diagonal; the parameter recovery section now explicitly reports their recovery quality.

(9) p 30: How are results affected by including the excluded participants? The level required to pass attention checks seems arbitrary. How was it chosen?

We have added a sensitivity analysis including the four excluded outliers, showing results remained qualitatively and statistically similar. With 10 binary attention checks, chance performance is 50%, meaning a participant scoring below 70% is performing only marginally above chance and likely not attending consistently. At the same time, 70% is permissive enough to retain participants who may have missed one or two checks due to momentary distraction.

(10) Figure F4: Typographically distinguish capital letters referring to panels in the figure from those referring to matrices.

Panel labels in Figure F4 are formatted in bold to distinguish them from italicised matrix notation.

(11) Table G1: Showing that differences between groups were non-significant before the intervention but significant after is not enough, you need to show that there was a significant interaction between time point and intervention. [I wrote this after reading the supplementary but before reading the main text. It turns out you know what I’m telling you here. You should mention it more prominently though, including in the abstract and the discussion, because in your chosen null-hypothesis significance testing framework, this is the crucial test of your study. I don’t think there’s any harm at all in being up-front about this - certainly much better than making excuses like the one about randomization at the top of page 11, which I recommend removing].

This is very right. The significant group × time interaction effects are now reported prominently in the main text Results and figure captions. We also wish to be transparent: the interaction tests were conducted post-hoc rather than as the primary analysis, which is the reverse of the correct order. In the interest of transparency we have left the structure of the results section intact rather than retrospectively reframing it.

Reviewer #3 (Recommendations for the Authors):

(1) I would encourage the authors to re-add some basic details regarding their power analyses from the supplement to the main text so the reader can immediately reference the intended effect size, which analysis/analyses were considered primary for the power analysis, etc.

More information about the power analysis have been added to the Participants section of the main text.

(2) Similarly, I wondered if there could be a little extra information on how the test-retest reliability was calculated (page 13) - on the first and last views of a video pre-intervention? I wasn’t sure - why do the authors only present ICCs for amusement/joy and disgust/horror? Seems useful to present all (particularly because there may be individual differences in habituation to some emotions).

We have expanded the test-retest section to clarify that reliability was assessed using duplicate videos shown three times pre-intervention, and now report ICCs with confidence intervals, Cronbach’s α, and habituation/sensitization tests for both disgust and amusement; the selection of these two categories reflects which videos were repeated in the design, they were chosen at random during experimental design.

(3) Regarding my comments about controllability/emotional control, I think the authors probably have two choices - address this head-on (e.g. with a note describing the relationship/distinction between these two concepts of controllability), or else avoid using it in one of the senses (I would suggest the mathematical sense since overriding the concept of cognitive control seems harder - the authors could use phrases like ’impact of emotional inputs’ instead of ’controllable’). In particular, the abstract could be clearer about the nature of controllability as implemented by the authors - this seems critical for communicability. Because of the high relevance of both of these ’control’ concepts to the paper, if the authors agree with my concern, I would also suggest changes throughout, such as in the results section phrasing: ”In those participants with high DERS-18 scores, the most controllable direction pointed towards disgust ( = 0.26, p = 0.006), and away from amusement ( = 0.26, p = 0.005) and calmness ( = 0.24, p = 0.011; though this did not survive Bonferroni correction).”

Thank you for this comment. Please see our response to the public review comments above, which we hope address this.

(4) Regarding the control intervention, which is great, is it possible the follow-up question/reminder affected the results – e.g., is there reason to believe that prospective regulation was the primary difference in the distancing group and not a retrospective effect via this question?

See our response to Reviewer 2 public review Comment 2 and the corresponding Limitations addition.

(5) Lastly, purely for interest, the authors could consider elaborating on their brief interpretation as to why difficulties in regulating emotions were specifically linked to the controllability of disgust, amusement, and calmness, but not other emotions (anxiety/sadness) (page 20). I wonder if there is a brief space to discuss the emotional specificity of these results further given the relevance to the wider literature on specific emotion types, e.g. fear vs disgust.

We have expanded the Discussion to elaborate on the emotional specificity of these findings. Amusement and disgust are strongly influenced by external events, suggesting that stimulus-driven controllability is particularly relevant for these emotions. By contrast, anxiety and sadness are maintained through internally generated processes such as rumination and anticipatory cognition, exhibiting greater emotional inertia over time, and their regulation may therefore be less sensitive to momentary stimulus controllability. This provides a mechanistic account of why controllability effects emerged selectively for disgust, amusement, and calmness, and aligns with growing evidence that emotion regulation is emotion-specific rather than domain-general.

Version 2

Abstract

Background

Emotion regulation strategies such as distancing are a core component of many evidence-based, effective psychotherapeutic interventions. They allow individuals to exert more ‘control’ over their emotional state. However, objectively disentangling how emotion regulation increases control has been difficult for reasons including a lack of a coherent theoretical framework for emotion control and insufficient experimental control over external inputs. Here, we apply a well-established theoretical framework for controllability to a tightly controlled experimental setup to examine the computational mechanisms by which emotion regulation interventions enhance emotional controllability.

Methods

109 participants were randomized to either a short emotion regulation intervention (distancing) or a control intervention. Both before and after the intervention, participants reported their emotional state along five dimensions repeatedly while watching a series of short, standardized, emotional video clips. A Kalman Filter was used to quantify how multidimensional emotional states changed with video inputs. The consequences of the emotion regulation intervention were examined by Bayesian model comparison, comparing models allowing for a change in intrinsic dynamics and/or input weights. Controllability was assessed using continuous measures derived from the controllability matrix, capturing relative differences in how emotional states can be driven by external inputs.

Results

The Kalman filter captured participants’ emotional trajectories, showing that emotional states were affected by the emotional videos; persisted; and interacted with each other. The distancing strategy made emotional states less externally controllable. It did so by altering two aspects of the dynamical system: by stabilizing specific emotional patterns and by reducing the impact of the external video clips.

Conclusions

Our study used a novel approach to examine emotion regulation, finding that a brief distancing intervention increased perceived emotion control by reducing how much external stimuli can control emotional states. This is due to both an increase in the intrinsic stability of certain emotional states; and a reduction of the sensitivity to certain extrinsic affective stimuli.

Introduction

Psychotherapeutic interventions are effective treatments for depression and anxiety, but the mechanisms of action responsible for these effects remain poorly understood (Kazdin, 2007; Carey et al., 2020). In addition, there is considerable variation between individuals regarding treatment response (Goldfried, 2013). Better targeting psychotherapeutic interventions according to individual needs or characteristics might improve therapy effectiveness. This requires understanding how existing psychological treatments work and for whom they work.

Previous research studying the mechanisms underlying psychological treatments has mostly focused on changes after the complete treatment courses (Klug et al., 2012). However, psychotherapeutic treatments are complex interventions involving different components (Luborsky et al., 2002), likely affecting different behavioural processes and acting via different mechanisms. As such, changes observed after the full course of treatment reflect the many many aspects of the intervention. In contrast, little research exists on how different components engage specific mechanisms.

In this study, we build on recent suggestions (Reiter et al., 2021; Wolpert et al., 2021; Holmes et al., 2018) and a promising new research direction (Brown et al., 2021; Huys et al., 2022; Dercon et al., 2023; Norbury et al., 2023), using cognitive computational techniques to study isolated interventions. This approach aims to better characterize and understand the change mechanisms of specific, key psychotherapeutic interventions. Specifically, we examine the effect of distancing—an emotion regulation technique—on the dynamics of emotions.

Emotion regulation strategies are an essential element of many psychotherapeutic treatments (Gross, 1998). Some emotion regulation strategies can alleviate symptoms of mood disorders and generally improve well-being (Berking et al., 2013; Boemo et al., 2022; Powers and LaBar, 2019; Somerville et al., 2022). Here, we define emotion regulation as using explicit strategies to intentionally up- or down-regulate positive or negative emotions. Effective strategies include problem-solving, reappraisal, acceptance, and distancing techniques (McRae and Gross, 2020; Webb et al., 2012). Distancing involves simulating a new perspective to increase the psychological distance from an event or situation and, with that, the emotional impact of a stimulus (Powers and LaBar, 2019; Webb et al., 2012). Distancing has been shown to reduce selfreported emotional experience reliably (Koenigsberg et al., 2009; Vrtička et al., 2012; Winecoff et al., 2011, 2013) and is associated with decreased amygdala activity even beyond the period of active regulation (Eippert et al., 2007; Walter et al., 2009; Domes et al., 2010). Distancing techniques are practical because they can be implemented in various situations with relatively low attentional demands and behavioural disruption (Powers and LaBar, 2019). We focus on distancing to gain insights into the mechanism of this specific treatment component in the hope of better understanding some of the complex processes underlying psychotherapeutic treatment effects.

Furthermore, emotion regulation research has mostly focused on the effects on individual emotions. Different emotions, however, are often related and influence each other. For instance, sadness can increase the likelihood of experiencing anger while emotions of different valence or arousal tend to inhibit each other: inducing happiness reduces experienced sadness, yet mixed states with both sadness and happiness can also occur. These interactions induce temporal dependencies between emotions, meaning that emotional states as a whole form a dynamical system fluctuating over time (Malamud and Huys, 2024; Durstewitz et al., 2020; Kuppens and Verduyn, 2017; Lange et al., 2022). As such, emotion regulation may be best conceptualized as the regulation of a dynamical system of emotions rather than as involving the regulation of individual emotions in isolation.

In recent years, network models have been developed to account for the dynamical systems properties of emotions (Borsboom and Cramer, 2013; Bos et al., 2017; Bringmann et al., 2013; Epskamp et al., 2018). This work has identified individual differences in affect dynamics which are linked to mood disorders (Bringmann et al., 2016; Kuppens et al., 2012; Pe et al., 2015, 2016; Sperry et al., 2020; Trull et al., 2015; Leemput et al., 2014). For example, increased inertia (temporal autocorrelation or how well an emotion can maintain itself) of negative affect has been identified in people with depression (Brose et al., 2015; Houben et al., 2015; Kuppens et al., 2010; Koval et al., 2012, 2013). Most interestingly, it has been suggested that there are discrete stable states (e.g. a depressed vs a happy state), so-called attractor states, between which people can transition (Durstewitz et al., 2020; Leemput et al., 2014; Wichers et al., 2016; Kuppens et al., 2012; Hosenfeld et al., 2015). Within an attractor state, the system is usually resistant to change, and under small disturbances (e.g. a stressful week), a person converges back to the current attractor state. A transition to another attractor state may occur if a perturbation is large enough or accumulated over time (Nelson et al., 2017). The transitions between states in a dynamical system are influenced by two important factors: the system’s intrinsic characteristics and its sensitivity to external driving forces.

Past work mostly focused on dynamical properties inferred from the emotion self-reports alone and has often neglected the role of external factors (Boemo et al., 2022; Malamud and Huys, 2024). This is a critical omission as the dynamical properties of a system are not fully identifiable unless the inputs are known — inputs can nearly arbitrarily alter the apparent dynamical system. For instance, persistent sad mood could be due to a constant external stimulus producing sadness rather than due to an internal persistence of sad mood. Ignoring the immediate context within which emotions fluctuate may hence lead to wrong conclusions about the underlying affective system. Indeed, it is well known that external stimuli profoundly impact the dynamics of emotional states, particularly in laboratory settings (Asutay et al., 2022; Rutledge et al., 2014; Villano et al., 2020; Vanhasbroeck et al., 2022).

Since both the natural changes in emotions over time and the impact of external factors on emotions are important, there is a compelling rationale to explore the potential effects of emotion regulation on these two key aspects of emotion processes. By doing so, we can gain valuable insights into the mechanisms underlying emotional regulation and better understand how emotions are controlled. Understanding which emotions can be most effectively changed through a certain techniques can provide clearer insights into the targeted effects of such interventions. Additionally, identifying reliable indicators of whether the emotional system is resistant or adaptable to changes could open up possibilities for creating personalized interventions.

Therefore, our aim was to examine the dynamic network effect of emotion regulation, specifically by considering an individual’s immediate context. We positioned emotion regulation within the framework of dynamical system theory and utilized tools from control theory to characterize the overall control properties of the emotional system. To briefly explain, a dynamical system is a series of linked differential equations (Brunton and Kutz, 2019; Durstewitz et al., 2020), each describing how one variable (here an emotion) changes over time. It permits emotions to influence each other and exhibits rich dynamical properties. The equations can incorporate the concept of control by allowing an input to drive one or more variables. Controllability, in this context, is a characteristic that depends on both how sensitive the system is to inputs and its inherent dynamics. It reflects how easy or hard it is to drive the system towards certain states. This definition differs from the common psychological use of emotion control or emotion regulation, which typically refers to the ability to deliberately manage, suppress, or modulate emotional responses in order to achieve desired emotional or behavioral outcomes (Gross, 1998, 2015). Importantly, high controllability in the control-theoretic sense does not imply tight regulation or reduced emotional variability; rather, it indicates that emotional states are more easily achieved through the selection of inputs. Likewise, sensitivity refers to the system’s responsiveness to inputs, and stability refers to whether emotional states return to a baseline following perturbations. For instance, if all emotions are positively linked to sad mood, then a happy state would be challenging to achieve, reducing the system’s controllability of single emotions. By modeling emotions this way, we can more precisely assess how a technique like distancing modulates the system: Either by changing its intrinsic dynamics (e.g., weakening feedback loops) or by altering how it responds to inputs (e.g., making emotional states more or less sensitive to external events).

In this study, we investigated how external emotionally-charged inputs influenced self-reported emotions over time (across multiple dimensions), and whether these inputs were essential to explaining the evolution of a rich, multidimensional emotional state. This enabled us to formally examine the impact of a distancing intervention on emotion dynamics whilst disentangling effects on intrinsic dynamics from alterations of external inputs. We also examined whether intervention-induced changes were moderated by measures of depressive or anxiety symptoms or difficulties in regulating emotions.

Methods

Participants

109 participants 18 years or older with current UK residence were recruited online on Prolific Academic (www.prolific.co). Prolific is widely used and reliable platform for online research due to its diverse participant pool and robust quality control measures (Peer et al., 2017; Palan and Schitter, 2018).

An a priori power analysis based on data from a pilot study (N = 40), aiming for 90% power to detect differences in the dominant unit vector of controllability, estimated a minimum sample size of N = 109 (cf. Supplementary Materials D Power Analysis).

Procedure and task

After indicating interest on the online recruitment platform, participants were forwarded to an electronic form of the participant information sheet. They could then provide electronic consent through an online form before being redirected to the experiment. The study duration was approximately 45 minutes, and participants were reimbursed £7.50/h through Prolific Academic after completion of study procedures.

In the experiment (Fig. 1), participants saw a sequence of short emotional video clips, each lasting 2-10 seconds and chosen for their efficacy in eliciting certain emotions. Video clips were from a previously-published database, with validated emotion ratings across multiple categories (cf. 2.2.2 Emotion-Inducing Stimuli; Cowen and Keltner 2017). After each video clip, participants reported their current emotional state in terms of two positive (amused and calm) and three negative emotions (disgusted, anxious, and sad). The selection of emotions was based on the specific affective qualities of the video stimuli provided in the validated database. While a traditional four-axis model typically plots emotions along dimensions of Valence (positivenegative) and Arousal (high-low/fast-slow), we selected five emotions to better capture the high-dimensional, nuanced states elicited by our stimuli. Emotions of disgust, anxiousness, and amusement were chosen as they were thought to be highly sensitive to video clip inputs. Conversely, sad and calm appear to be more persistent emotions which we expected to be informative about stability. Participants were instructed to use a slider to indicate how strongly they felt each of the emotions at that moment in time, with options ranging from “not at all” to “very”. Participants had 30 seconds to report their emotional state. If they did not manage to rate all emotions within 30 seconds, the experiment moved on to the next video without their completed rating.

Task Description.

The large arrow at the top displays the course of the experiment. Both video blocks comprised 54 video clips (2-10 seconds), and after each video clip, participants had 30 seconds to rate their emotional experience based on five emotions. Both blocks also had a baseline emotion rating before the videos started. Participants underwent either a distancing (emotion regulation) or a relaxation (control) intervention between the two video blocks. After the second block of video clips, participants completed three standardized psychological questionnaires measuring symptoms of depression (PHQ-9) and anxiety (GAD-7) and emotion regulation difficulty (DERS-18). PHQ-9: Patient Health Questionnaire. GAD-7: Generalized Anxiety Disorder Assessment. DERS-18: Difficulties in Emotion Regulation Scale.

After watching the first block of 54 video clips, participants were randomized to undergo either a distancing (emotion regulation) or a relaxation (control) intervention (cf. 2.2.1 Intervention). Following the intervention, participants watched a second block of 54 video clips and again rated their emotions after each video. A different set of video clips was shown, but the sequence of emotions targeted was matched to the first block.

Intervention

The emotion regulation intervention was based on a distancing appraisal strategy, i.e. “Leaves on a Stream” (adapted from Hayes et al. 2006). This technique involves viewing emotions and thoughts as events passing through one’s mind rather than getting “sucked in” by them. The script employed a visualization strategy, instructing participants to imagine they were standing by a stream with leaves floating gently past them. They were then told: “When an emotion or thought comes up, imagine you place the thought on one of those leaves and that you are watching the leaf - carrying your emotion or thought - float away, disappearing behind a corner or in the distance.” In the control intervention, participants were asked to engage in a relaxation exercise. They were told the same storyline as in the distancing intervention but without connecting the stream and leaves to their emotions and thoughts (the full text of both interventions is reproduced in the Supplementary Materials A Intervention Text).

Furthermore, in the group allocated to the distancing intervention, the phrasing of the question about their feelings in the second video block reminded participants about the intervention, stating: “You observed your emotions and let them pass like the leaves floating by on the stream.” This was not the case for the control group.

To reduce demand effects, the instructions were framed to suggest that the experiment was attempting to understand whether the distancing was helpful or not, i.e. participants were explicitly encouraged to report that the intervention was not helpful if they did not perceive it as effective.

Emotion-Inducing Stimuli

The emotion eliciting video clips stem from a validated database (Cowen and Keltner, 2017) (link to video database: https://s3-us-west-l.amazonaws.com/emogifs/uncensored.html). This study analyzed emotional responses to 2,185 short videos, identifying 27 distinct emotional experiences categorized through a survey completed by 853 English-speaking US participants. The videos were collected from online sources depicting diverse emotionally significant situations, including cute animals, natural landscapes, distressing scenes such as feces and vomit, accidents and dangerous stunts, and many others.

For our experiment, we selected video clips that fit into five target emotion categories: Amusement/Joy, Disgust/Horror, Sadness/Sympathy, Calmness/Aesthetic Adoration, and Anxiety/Fear. We chose clips that showed a high mean and a low entropy in the ratings from Cowen and Keltner (2017) for these emotion categories, to ensure they would specifically evoke the intended emotion. We identified 20 videos from each of the five categories, resulting in a total of 100 videos. The ratings from Cowen and Keltner (2017) for each chosen video and the mean ratings over videos from an emotion category are shown in Supplementary Materials B Video Clips Fig. B.2, respectively Fig. B.1.

Study Sequence

In total, each participant viewed 54 video clips before and 54 video clips after the intervention. The video order was pseudorandomized for the first half. In addition, two randomly chosen videos were repeated three times within the first block to investigate the reliability of the ratings. The second block of videos contained different video clips, but the sequence of target emotions was the same as in the first block (including the repeated videos). All participants saw the identical video clip sequences, but before and after intervention sequences were counterbalanced across participants. Participants provided 110 ratings in total (55 before and 55 after the intervention) including a baseline rating before the first video of each block. The length of the experiment was based on simulation and recovery work (cf. Supplementary Materials C Experimental Design).

Self-Report Measures

At the end, participants were asked to complete three self-report questionnaires. To assess depressive and anxiety symptoms, we used the Patient Health Questionnaire (PHQ-9; Kroenke et al. 2001) and the Generalized Anxiety Disorder Assessment (GAD-7; Spitzer et al. 2006), respectively. In addition, we used a short version of the Difficulties in Emotion Regulation Scale (DERS-18; Victor and Klonsky 2016) to assess participants’ ability to identify, accept, and manage their emotional experiences. The questionnaires were conducted after the experiment to prevent them from affecting participants’ behaviour during the study.

Attention Checks

To maintain and monitor attention, participants were asked to detect a black cross, which could be shown briefly before the video clip started. 10 such attention checks were included, 4 in the first block and 6 in the second. Participants with an accuracy of less than 70% on the attention checks were excluded.

Computational Modelling

Conceptual Definitions

During the experiment, we asked participants to report their emotions after watching each video, resulting in a sequence of emotional state reports. To capture the temporal structure of this data, we used a linear Gaussian state space model, specifically the time-invariant Kalman Filter (Durbin and Koopman, 2012; Roweis and Ghahramani, 1999). This model is simple and tractable, accounting for possible errors in participants’ self-reports and allows for external inputs that might influence their emotions. Essentially, this model helps us track the true emotional process as it evolves based on the participants’ reports.

The Kalman Filter suggests an underlying latent process, referred to as the true emotional process, evolving in response to a sequence of emotional state reports xtt=1T conditional on video inputs utt=1T (Fig. 2A).

The latent (unobserved) emotional state vector zt comprises the activation of each emotion category zti at time t. These latent emotions are assumed to evolve according to a discrete, linear first-order Markov process. The matrix A defines the dynamics matrix, which captures the degree to which an emotion carries over from one time point to the next and the degree to which it predicts future emotions. The bias term h relates to the steady-state baseline of the emotional system. In a dynamical system without external inputs, emotions might not naturally decay to zero; instead, they may converge toward a stable mean rating z0 = (I - A)−1h. The matrix C captures the impact of videos, with the entry Cij capturing the impact of a video of category j on emotion category i. The video category shown at each time point was identified through one-hot labelling in the binary input vector ut i.e. each video was identified in the vector u by the main emotion it was targeting. Rather than using fixed average ratings from the stimulus database, we utilized one-hot encoding to allow the input weight matrix C to directly estimate the unique intensity and sensitivity for each participant. The subjects’ ratings [xt]i on each category i were mapped directly onto the latent emotion states [zt]i. Critically, however, Gaussian noise was added to both the latent temporal evolution and the observation processes with separate diagonal covariances Σ and Γ, respectively. This is one of the key differences to standard emotional time-series analyses (Borsboom et al., 2021), in that this explicitly allows for noise or errors in the emotion ratings, and this noise can be ‘smoothed over’ through the filtering process.

Dynamical System and Controllability.

A) shows a graph visualization of the linear dynamical model, including external inputs (ut; emotional video clips). zt describes the latent (unobserved) emotion states evolving based on a Markov process and directly mapping onto the acquired emotion ratings xt. B) shows the trajectories of a two-dimensional system (ratings of “amused” and “anxious”) starting from a randomly chosen initial point without external inputs. Whereas anxiety decays independently of amusement, amusement is influenced by anxiety and thus, the trajectory of amusement is more complex and does not simply exponentially decay to zero. However, the more anxiety decays, the more the influence of anxiety on amusement decreases; and both variables converge towards zero. C) shows the trajectory of both emotion ratings plotted against each other. The blue x indicates the starting point. D) displays the independently evolving trajectories of the transformed variables z˜ resulting from the projection of the state variables z onto the eigenvectors of the dynamics matrix A. E) The left singular vectors V of the controllability matrix define an energy ellipsoid where the left singular directions corresponding to higher singular values (S) are more controllable and vice versa. That means with the same effort one can go further into the most controllable direction (v1) and least far into the least controllable direction (v3)

Model Comparison

To understand the dynamics of the emotion reports elicited by the video sequences, we asked whether the different components of the Kalman Filter were indeed necessary to provide a parsimonious account of the observed data. To do this, we built increasingly complex models. The simplest model included only Gaussian noise, i.e. it assumed emotion ratings varied randomly over time. The next model additionally contained either a dynamics matrix (A), an input weight matrix (C) or both. The most complex model included a dynamics matrix (A), an input weight matrix (C) and a bias (h). Additionally, we examined variations of these models where we constrained the dynamics matrix (A) and/or the input matrix C to be diagonal. Finally, models were fitted separately to each individual’s emotion time-series using the python package Pykalman: https://pykalman.github.io/.

We calculated the Bayesian Information Criterion (BIC; Schwarz 1978) for each individual for each model based on an individual’s model loglikelihood by penalizing for the number of parameters in the model. Models were then compared using the BIC at the group level. Cumulative model weights (ωj), normalize raw BIC scores so they can be interpreted as the relative probability that a specific model is the best one among the set being compared: ωj=e−ΔBICj∑i=1Ne−ΔBICj where ΔBICi = BICi – min(BIC), j indicates the most parsimonious model and N the number of models (Kass and Raftery, 1995; Burnham and Anderson, 2004). A post-hoc random-effects Bayesian model selection (Rigoux et al., 2014) suggests that the group-level effects observed reflect some heterogeneity across participants (c.f. Supplementary Materials G Random-Effects Bayesian Model Selection.)

Stability

After extracting the parameters of the most parsimonious model, we investigated the eigenstructure of the dynamical system. Briefly, a linear dynamical system where variables interact, like the Kalman Filter model, can be decomposed into separate systems of non-interacting variables (Durbin and Koopman, 2012; Strogatz, 2015). This is achieved through an eigendecomposition of the dynamics matrix. By projecting the vector of state variables z on each of the eigenvectors, new combined state variables z˜ can be defined. These eigenmodes evolve independently, i.e. no longer interact (Strogatz, 2015) and identify the effective emotional combinations which determine the evolution of an individual’s affective state. The eigenvector corresponding to the largest eigenvalue of the dynamics matrix A identifies the most stable combination of emotions, provided the eigenvalue is within the unit circle (|λ| < 1), indicating a system that converges to a steady state. In contrast, the eigenvector corresponding to the smallest eigenvalue identifies the combination of emotions that is most transitory and least persistent (for an example cf. Fig. 2B-D).

Controllability

Next, we investigated the controllability of the dynamical system. A system is more controllable if smaller inputs u are required to move its state z to any required value (Brunton and Kutz, 2019). We computed the controllability matrix (𝒞) for each participant as follows: 𝒞=[C AC A2C … An−1C] (3)

The controllability matrix (𝒞) combines the dynamics matrix (A) and the weights of the external input (C) to the dynamical system. If the rank of this controllability matrix is equal to the system’s dimension, the system is controllable (Brunton and Kutz, 2019). A controllable system means that any state z can be achieved through the appropriate choice of external inputs u. How controllable the system is captured by the strength of input |u| required. We investigated the characteristics of the controllability matrix using singular value decomposition. The left singular vectors of the controllability matrix define an energy ellipsoid (Fig. 2E). The left singular vectors corresponding to higher singular values identify the more controllable directions in the state space and vice versa. The more controllable a direction is, the less input energy is required to steer the system in that specific direction. In other words, an input of a given strength |u| can move the system further in a direction which aligns with a more controllable direction than a less controllable one (Brunton and Kutz, 2019).

Statistical Analysis

We used one-sided two-sample t-tests to test whether the emotion ratings averaged over video clips from the same video category for the emotion which was aimed to be elicited were higher than for the other emotions. In addition, two-sided one-sample t-tests were performed to test whether the mean emotion ratings, autocorrelation coefficients, and cross-correlation coefficients of emotion time-series significantly differed from zero. Those tests were all conducted on data before the intervention.

To investigate intervention effects, the principal analyses was a two-sample Hotelling T2 tests to compare multivariate variables (e.g. eigendirections) between groups after the intervention. The randomized group allocation allowed us to focus on potential effects after the intervention. We also performed multivariate ANOVAs (MANOVAs) to test for an interaction effect between time (before and after intervention) and group. To zoom into single emotions, we used non-parametric tests because most dynamical and controllability features were not normally distributed. Mann-Whitney U tests were conducted to compare emotion variables, such as mean emotion ratings and eigenvector directions, between the intervention groups and one-sample Wilcoxon signed-rank test to compare variables before and after the intervention within an intervention group.

Multiple linear regressions were performed to investigate associations between emotion dynamics (dependent variable; DV) and symptoms (independent variable; IV) controlling for the intervention group (G): DV = β0 + β1IV + β2G. Finally, to investigate whether psychological wellbeing moderated the effect of the intervention on emotion ratings, we examined the interaction effect between symptom score and intervention group: DV = β0 + β1IV + β2G + β3(IV × G) + ϵ. All variables were z-scored for the regressions. For all above-mentioned analysis types, we used Bonferroni-correction to correct for the number of conducted tests.

Results

Participants

109 participants completed the experiment, but one participant was excluded due to attention check failure. Analyses hence focused on, 108 participants (57 randomized to the distancing intervention).

Eliciting Complex Emotional States with Videos

In the first video block before the intervention, each video induced the dominant (targeted) emotion as intended (Fig. 3A). Changes in the dominant emotion ratings for each video were significantly higher than changes in other emotions (t ∈ [5.9,26.2], all p < 0.001; Fig. 3B and Supplementary Material Table K.4). This shows that the sequence of emotional videos reliably induced the targeted emotions. The correlation between our participants’ emotion ratings and those reported by Cowen and Keltner (2017) was moderate to strong across all emotion categories, with correlations of r = 0.74 for Disgust/Horror, r = 0.65 for Amusement/Joy, r = 0.5 for Calmness/Aesthetic Adoration, r = 0.71 for Anxiety/Fear, and r = 0.6 for Sadness/Sympathy (all p < 0.001). Thus, we replicated the ratings from Cowen and Keltner (2017) reasonably well.

Emotion Ratings.

A) The heatmap shows for each emotion eliciting video category the ratings from Cowen and Keltner (2017) averaged over the two emotion categories of interest (left y-axis). The coloured lines with dots report the emotion ratings from our experiment for each emotion after watching a video clip averaged over participants before the intervention occurred for the randomized and matched video sequences separately (right y-axis). B) shows the change between the rating (t) after a certain video and the previous rating (t - 1) averaged over participants and all trials, including videos from the same video category. Subplot B has the same emotion categories as subplot A (left label of A). The bar represents the mean of the change in ratings, the black line shows the standard deviation, and the * indicates a significant difference from zero. The black frame shows the dominant emotion for the specific emotion category (the emotion intended to be elicited by watching videos from that category). The unframed bars show that videos from a specific emotion category also affected non-dominant emotions. C) shows the autocorrelation coefficient averaged over participants for five lags for each emotion. The line indicates the mean and the shaded area standard deviation over participants. D) shows the mean and standard deviation of the cross-correlation coefficients between emotion time-series averaged over participants. Significance *≤ 0.5, **≤ 0.1, ***≤ 0.001, ****≤ 0.0001

Furthermore, changes in emotions other than the dominant one within a video category were significantly different from zero (|t| ∈ [5.1,19.3], all p < 0.001; Fig. 3B and Supplementary Material Table K.4). This indicates that the videos induced complex and multi-faceted emotional states. We also observed that emotional responses had a dynamic component. This was evident in the autocorrelation of each emotion rating, where an effect of the previous time-point is apparent during the rating of the next video (AR(1) ∈ [0.07 – 0.25], all p ≤ 0.001; Fig. 3C and Supplementary Material Table K.5).

Dynamic effects can also be observed in the cross-correlation between emotion time-series showing that emotions interact with each other (|r| ∈ [0.28 – 0.52], all p < 0.001; Fig. 3D and Supplementary Material Table K.6).

Finally, test-retest reliability was assessed using duplicate videos shown three times pre-intervention for disgust and amusement. Amusement ratings demonstrated excellent reliability (video sequence 1: ICC(2,1) = 0.79, CI = [0.69,0.86], p < 0.001, α = 0.92, video sequence 2: ICC(2,1) = 0.78, CI = [0.670.86], p < 0.001, α = 0.91) with no significant habituation (p > 0.05). Disgust ratings showed moderate-to-good reliability (video sequence 1: ICC(2,1) = 0.59, CI = [0.43,0.72], p < 0.001, α = 0.84 , video sequence 2: ICC(2,1) = 0.78, CI = [0.670.86], p < 0.001, α = 0.91), though significant sensitization effects (p < 0.001 and p = 0.018), with ratings increasing across repeated viewings (MAD (average absolute change in ratings between the 1st and 3rd viewing) = 16 and MAD = 24).

In summary, the experimental setup resulted in reliable elicitation of complex, high-dimensional emotional states, allowing for the joint characterization of emotional input sensitivity and emotion dynamics.

Establishing a Dynamical Model

We next examined the dynamical properties of the emotion ratings, and the interaction with the emotion inputs using a Kalman Filter. First, we compared different models based on the group-level BIC to evaluate which dynamical components are required to capture the data over the whole experiment, i.e. the concatenated time-series before and after the intervention (Fig. 4A). We found that the most parsimonious model, carrying 99% of the cumulative model weight, included a dynamics matrix A, a full input weight matrix C and diagonal noise covariances Σ and Γ. Importantly, data generated from this model accurately captured the observed data sequences as shown in the blue-shaded part in Figure 4C, with the model explaining on average 60% of variance in emotional ratings (mean R2 = 0.56 – 0.64 across emotions) and a mean temporal correlation of r = 0.74 – 0.80, confirming that both the timing and magnitude of emotional responses were well reproduced. Hence both external inputs and intrinsic dynamics are required to explain the self-reported emotion ratings. The recoverability of the matrices A and C in the most parsimonious model is shown in Supplementary Materials F Parameter Recovery.

Model Evaluation.

A) shows the differences in Bayesian Information Criterion (BIC) scores for all models tested compared to the most parsimonious model (blue star). All models were separately fitted to individuals’ emotion rating time-series over the whole experiment. The left y-axis shows the number of free parameters for each model. The most parsimonious model included a full dynamics matrix, full input weight matrix and diagonal noise covariances for observation and process noise. B) shows differences in BIC scores of models, allowing for parameters to change after the intervention. While in the control group, a model in which all parameters stayed the same best explained the data (grey star), in the distancing group dynamics matrix, input weights and noise covariances differed in the most parsimonious model (purple star). C) shows empirical (blue) and simulated data (orange) from two randomly selected participants. One of those was allocated to the control intervention and one to the distancing intervention. We simulated data from a linear state space model using the parameter estimates derived from fitting the Kalman Filter to individuals’ emotion rating time-series. The vertical grey dashed line indicates that a video was shown at that time-point which stems from the category aiming to elicit that specific emotion. Blue shading indicates the period before, and green shading after the intervention. Data was simulated based on different parameter sets before and after the intervention.

Effects of Distancing on Emotional Responses

We first examined the effect of the intervention on the multivariate reported emotions averaged over trials (Fig. 5A). The mean ratings averaged over the video block after the intervention were significantly different between the distancing and the control group (T2 = 24.48, F = 4.71, p < 0.001). In contrast, before the intervention the mean ratings did not show a significant difference between groups (T2 = 6.53, F = 1.26, p = 0.29). The within-subject change in average ratings (after minus before intervention) differed significantly between groups (T2 = 52.66, F = 10.14, p < 0.001). Additionally, we observed a significant interaction between time and intervention group (F(5,208) = 2.6, p = 0.026).

Intervention Effects.

A) shows the average ratings separated in before and after the intervention and for both intervention groups. Purple shades refer to the distancing group and grey shades to the control group. The boxes show the quartiles of the ratings, while the whiskers extend to show the rest of the distribution. The horizontal bars in the boxes indicate the median, and the green triangles the mean across participants. B) shows mean and standard deviation for the elements of the input weight matrix (C matrix) averaged across participants before the intervention occurred. C) shows the effect of distancing on the input weight matrix (C matrix), which is computed as the mean change (after minus before intervention) divided by its standard deviation for each matrix element only in participants allocated to the distancing intervention. D) The boxplot presents the quartiles of the dominant eigenvalues of the dynamics matrix, while the whiskers display the rest of the distribution. E) The boxplot illustrates the quartiles of the dominant singular values of the controllability matrix, with the whiskers showing the rest of the distribution. F) The polar plots show the left singular vector direction of the dominant singular value separated between participants in the distancing (purple shades) and the control (grey shades) group before (left) and after (right) intervention. Significance *≤ 0.05, **≤ 0.01, ***≤ 0.001, ****≤ 0.0001. We report the significance after Bonferroni correction for testing five different emotions. Interaction effects (time × group) were significant for mean emotion ratings (F(5,208) = 2.6,p = 0.026) and the most controllable direction (F(5,200) = 2.5,p = 0.03), panels A) and F) respectively.

At the group-level, all emotion ratings were significantly reduced after the distancing intervention (U ∈ [688,910] for all comparisons, all p ≤ 0.001), except for ratings of calmness which were increased (U = 1824, p = 0.02; though this does not survive Bonferroni correction p≤0.055≤0.01. Furthermore, while this pattern was observable in the distancing group (W ∈ [68,308] for all comparisons, all p < 0.001, before vs after), no changes were detectable in the control group (W ∈ [401,561] for all comparisons, all p ≥ 0.01, except for amused W = 273.0, p < 0.001).

Distancing also affected the temporal variability in emotion ratings (group comparison after intervention: T2 = 36.53, F = 7.03, p < 0.001). The variances of all emotions were reduced (U ∈ [576,983] for all comparisons, all p ≤ 0.004) in the distancing group compared to the control group and they significantly differed before and after the intervention within the distancing group (W ∈ [133,246] for all comparisons, all p < 0.001). However, the interaction effect in a MANOVA was not significant (F(5,208) = 1.72, p = 0.13).

Overall, the distancing intervention strongly affected the mean emotions subjects reported and the variability of emotions (cf. full table in Supplementary Materials Table K.7).

To examine whether the observed group differences are attributable to a demand effect, we analyzed the time taken by participants to rate their emotions on the sliders. We reasoned that if a demand effect were present, participants would respond more quickly as they would already have decided in advance of the question how to respond. Using a mixed ANOVA with time (before vs after the intervention) and intervention groups as factors, we found evidence of a significant interaction effect between time and group (F(1, 106) = 6.46, p = 0.01). However, interestingly, the control group demonstrated a decrease in reaction time (before: M = 12029, SD = 2813 (ms); after: M = 10964, SD = 2713 (ms); t = 5.87, p < 0.001), while the distancing group did not show a significant difference in reaction time before and after the intervention (before: M = 11876, SD = 2796 (ms); after: M = 11518, SD = 2896 (ms), t = 1.73, p = 0.09).

Effects of Distancing on Dynamics of Emotions

Next, we examined whether the intrinsic dynamics or the input weights or both were altered by the distancing intervention (Fig. 4C). In principle, the observed effect of distancing on the mean and variance of emotion ratings can be explained either by changing the internal dynamics or by changing the input weights alone (cf. Supplementary Material Figure H.6). We hence applied model comparison to examine whether either a change in the dynamics, or in the input weights, or in both was required to capture the data. In the control group, the most parsimonious model (99% of model weight) was the one where the dynamics before and after the intervention stayed the same. That is, there was no evidence for a change in either input weights or dynamics. By contrast, in the distancing group, a model where noise, dynamics, and input weights changed with the intervention provided the most parsimonious account of the data (99% of model weight). Again, data generated from this model was able to capture the observed data sequences accurately (Fig. 4D). Overall, this suggests that the distancing intervention changed both the dynamics and the influence of the videos.

We then examined the specific dynamical features that changed in response to the distancing intervention. Four participants had to be excluded as they were outliers in dynamical characteristics (cf. Supplementary Materials E Exclusion). The following analyses were hence based on 104 participants (54 randomized to the distancing intervention). A sensitivity analysis including the four subjects previously excluded due to input weight outliers showed that the results remained qualitatively and statistically similar.

Distancing altered the input weights (C matrix) shown in a group comparison of the matrix norm (after intervention: distancing group M = 256, SD = 218; control group M = 400, SD = 222; U = 1897, p < 0.001) and in Fig. 5B&C. Input matrices averaged across subjects for both groups at pre- and post-intervention, along with their difference, are shown in Supplementary Material Figure J.9.

Furthermore, a linear dynamical system can be decomposed into eigenmodes – parallel, independent dynamical systems – using an eigendecomposition of the dynamics matrix A (cf. 2.3.3 Stability). Distancing altered the composition of the most stable emotional eigenvector, the one corresponding to the largest/dominant eigenvalue of the dynamics matrix A(T2 = 14.81, F = 2.85, p = 0.019; Fig. 5D). This component was also more stable (decayed more slowly) in the distancing group (shown by a higher dominant eigenvalue: distancing group M = 0.69, SD = 0.21; control group M = 0.52, SD = 0.22; U = 762,p < 0.001; Fig. 5D) after the intervention. Hence, when controlling for the emotional input, the distancing intervention had specific effects on how different emotional states persisted and interacted.

A general analysis of the dynamics matrix A revealed that while calm exhibited the highest persistence, consistent with our hypothesis (0.59 ± 0.85), sadness did not show the expected stability (0.09 ±0.58), and other emotions demonstrated low persistence (0.13 to 0.24) alongside substantial individual variability (cf. Supplementary Material Figure J.8).

Effects of Distancing on Controllability of Emotions

The dynamics matrix A and the input weight matrix C jointly determine the extent to which the emotional state can be controlled by external inputs. This can be formally assessed through a measure called controllability, which we turn to next (cf. 2.3.4 Controllability). Controllability formalizes how strong the inputs to the system have to be to move the dynamical system around, i.e. how ‘reactive’ the dynamical system is to inputs.

The intervention altered controllability overall (group comparison of matrix norm after intervention: distancing group M = 335, SD = 284; control group M = 509, SD = 292; U = 1851, p = 0.001).

Controllability can, however, vary, with the system being more controllable in certain directions than others. This can be examined by studying the eigenspace of the controllability matrix 𝒞, which is based on a combination of dynamics and input matrices, A and C. The eigenspace corresponding to the dominant eigenvalue of the controllability matrix, describing the most controllable direction, differed between the intervention groups after (T2 = 24.67, F = 4.74, p < 0.001), but not before (T2 = 3.69, F = 0.71, p = 0.618), the intervention. A posthoc MANOVA revealed a significant interaction between time and intervention group on the most controllable direction (F(5,200) = 2.5, p = 0.03). Distancing altered which combination of emotions was most controllable (Fig. 5F), with combinations involving more calm (U = 874, p — = 0.002) and less sad emotions (U — = 1926, p < 0.001) being most controllable. This direction was, overall, less controllable (lower singular value; U = 1844, p = 0.001; Fig. 5E) in the distancing group.

Psychopathological Correlates

The responsiveness to external influences is thought to be altered in mental illness. Hence, we examined controllability in relation to the three assessed psychological questionnaires, focusing on the self-reported emotion regulation difficulties. All questionnaires showed a good internal consistency (Cα ∈ [0.89,0.91]). The total scores of the three questionnaires were highly correlated (r ∈ [0.69,0.75], p < 0.001) and distributions of scores were skewed towards lower values (cf. Supplementary Material Figure 1.7).

Psychological well-being was significantly related to emotion ratings at baseline (t0), emotion ratings averaged over the first video block (t1) and the second video block (t2). Ratings of disgust, sadness and anxiousness were positively associated with PHQ-9, GAD-7, and DERS-18 total scores (β ∈ [0.18,0.56] for all comparisons, all p ≤ 0.03; except for GAD-7 and disgust at t0 and t2 β < 0.12, p > 0.2). Ratings of calmness decreased with increasing symptoms (β ∈ [–0.18, –0.4] for all comparisons, all p ≤ 0.05). There was no relationship between amusement and symptoms. However, critically none of the symptom scores interacted with group allocation in predicting change in emotion ratings (all p > 0.05; cf. Supplementary Material Table K.9).

Psychopathological symptoms were related to aspects of the dynamics of emotions. Difficulties in distancing as measured by the DERS-18 were related to the most controllable direction in the emotion space. In those participants with high DERS-18 scores, the most controllable direction pointed towards disgust (β = 0.26, p = 0.006), and away from amusement (β = –0.26, p = 0.005) and calmness (β = –0.24, p = 0.011; though this did not survive Bonferroni correction). In other words, self-reported emotion regulation difficulties were related to needing less effort to drive disgust and more effort to steer the emotional experience towards calm and amused. The effect of DERS-18 total score on disgust was even more prominent after the intervention (β = 0.35, p < 0.001).

There was no evidence that the symptoms moderated the effect of distancing (all p > 0.05; cf. Supplementary Material Table K.10 & K.11).

Discussion

Psychotherapeutic approaches effectively treat a variety of psychiatric conditions. However, the mechanisms underlying these psychotherapeutic effects remain unclear (Kazdin, 2009). By isolating a specific intervention component, such as emotion regulation, we aim to gain a deeper understanding of the processes involved in this specific treatment effect. Emotion regulation techniques are core components of psychotherapy (McRae and Gross, 2020; Powers and LaBar, 2019; Gross, 2015). To study emotion regulation strategies, it is essential to investigate emotions and consider that they fluctuate in intensity and frequency over time, potentially giving rise to a complex dynamical system. We believe that by examining the emotion dynamics of a particular individual, we can better understand how emotions change and how they respond to different interventions.

Here, we studied the impact of an emotion regulation technique called distancing on the dynamics of emotions over time. We used rich and powerful video stimuli to evoke and alter a complex emotional state in participants. This approach allowed us to achieve three main objectives: first, we established how different emotions influence each other over time in a dynamic system; second, we assessed the relative impact of the video clips on these emotions; and finally, we disentangled the effect of distancing on the stability and controllability of these emotional states.

Importance of external inputs in a dynamical system

Our findings revealed that video clips evoked complex patterns in self-reported emotional experiences, including interrelated emotional trajectories. This highlights the significance of considering multidimensional emotional experiences, their interactions, and temporal dependencies, rather than focusing on a single emotion in isolation (Kendler, 2016; Hitchcock et al., 2022; Durstewitz et al., 2020; Kuppens and Verduyn, 2017; Lange et al., 2022). Computational modeling results suggest that external inputs (video clips) were important in explaining a higher dimensional emotional state. In other words, models that included these inputs provided a better explanation of the self-reported emotion responses. This is in line with recent experimental research demonstrating how unexpected and personally-relevant events are associated with measurable changes in the time course of individuals’ emotional responses (Villano et al., 2020; Rutledge et al., 2014; Eldar et al., 2016; Asutay et al., 2022; Lapate and Heller, 2020). Omitting inputs when studying the properties of a dynamical system can lead to inaccurate conclusions as the inputs can mask or alter the apparent dynamics of the system. For instance, Vanhasbroeck et al. (2022) has observed that nonlinearity observed in affective time series in some individuals was the result of external inputs rather than underlying nonlinearity in affect. Hence, to ensure an accurate understanding of the affective system, it is crucial to consider the relationship between affect dynamics and the immediate environment.

While emotional self-reports are inherently subjective, we note that the regularities governing subjective emotional experience are of primary scientific interest in their own right, and no consensus definition of a ‘genuine’ emotional state that supersedes self-report currently exists.

Distancing effect on emotional controllability

First, the distancing intervention reliably reduced ratings and variability of both positive and negative emotions but increased ratings of a specific variable, namely feeling calm.

Second, the modeling approach enabled a formal, quantitative assessment of the impact of the distancing intervention on these components. Model comparison showed that the combination of dynamic and controllability signatures changed due to the distancing intervention. Specifically, a brief distancing intervention stabilized intrinsic emotion dynamics and reduced the impact of external stimuli. This finding is significant because static mean differences could be attributed to changes in either dynamics or input sensitivity alone. Interestingly, the intervention affected emotions differently: the controllability of calmness decreased, while sadness increased relative to other emotions. This suggests that some emotions may be more responsive to regulation through distancing. Here, controllability refers to the formal control-theoretic notion of state reachability given external inputs, rather than deliberate or effortful emotion regulation in the psychological sense.

Emotional controllability and psychopathology

Psychopathology has been associated with deficits in cognitive control (Grahek et al., 2018; Snyder, 2013). The belief that emotions can be controlled to some extent was linked to a decrease in symptoms of anxiety and depression. This connection can be attributed, at least in part, to the fact that individuals with such beliefs tend to employ adaptive strategies for regulating their emotions more frequently (Somerville et al., 2022).

Distancing involves several cognitive control processes, such as taking a step back from a situation, observing it objectively, and cognitively re-framing the experience to focus on positive aspects rather than negative feelings (McRae and Gross, 2020; Ochsner et al., 2004; Dorfel et al., 2014; Staudinger et al., 2009). These processes may include inhibiting pre-potent evaluations, shifting attention away from external stimuli towards the self, and maintaining an intention to detach (McRae and Gross, 2020; Ochsner et al., 2004; Dorfel et al., 2014; Staudinger et al., 2009). Our study demonstrated that the distancing intervention effectively reduced emotional variability and increased feelings of calmness, consistent with improved emotion regulation at the behavioral level. Furthermore, our modeling showed that the distancing intervention stabilized intrinsic emotion dynamics and reduced the impact of external stimuli. Importantly, these changes reflect alterations in system dynamics and input sensitivity, rather than improvements in deliberate emotion control per se. This suggests that distancing may promote emotional stability by changing how strongly emotional states are driven by external events, rather than by increasing effortful control over emotions.

This is consistent with evidence that individuals with high self-control are more successful in regulating emotions (Paschke et al., 2016) and that emotional stability is linked to self-control (Daly et al., 2014; Tangney et al., 2004; Oaten and Cheng, 2006).

An interesting consideration is that cognitive biases may influence emotion regulation ability, thereby contributing to sustained negative affect and reduced levels of positive affect (Joormann and Tanovic, 2015; McRae et al., 2012). For instance, individuals with depression often have difficulty inhibiting negative thoughts or shifting their attention away from negative stimuli. This can lead to a cycle of negative thinking that worsen symptoms (Beck and Haigh, 2014). Our study found that difficulties in regulating emotions were specifically linked to the controllability of emotions such as disgust, amusement, and calmness, indicating that these emotions may be more intrinsically tied to individuals’ perceived ability to regulate their emotions (Gross, 2015; Mitchell, 2021). Notably, emotions like amusement and disgust are strongly influenced by external events (Rozin and Fallon, 1987; Gross and Levenson, 1995), suggesting that the emotional impact of external stimuli plays a critical role in the complexity of emotion regulation difficulties in psychopathology. In contrast, anxiety and sadness are maintained through internally generated processes such as rumination and anticipatory cognition, exhibiting greater emotional inertia over time (Kuppens et al., 2010). Regulation of these emotions may therefore be less sensitive to momentary stimulus controllability, potentially explaining why controllability effects emerged selectively for disgust, amusement, and calmness. These findings align with growing evidence that emotion regulation is emotion-specific rather than domain-general (Sheppes et al., 2015). However, contrary to our expectation, we did not find a correlation between symptoms of psychopathology and the efficacy of the distancing intervention.

Future directions

Emotion dynamics and their controllability are complex and context-dependent. Future research need to examine whether the findings discovered in this study can be applied beyond the realm of video-induced emotions. The key challenge here will be the appropriate, but necessary, characterisation of inputs. Unless inputs are well-known, the true internal dynamics are not identifiable. However, in more naturalistic environments this appears very difficult (Malamud and Huys, 2024).

Limitations

This study comes with several limitations. First, it is challenging to identify the degree of influence of possible demand effects. We attempted to avoid this influence by providing participants with the sense that distancing does not work for everyone, and we were interested in how it works for them; however, this might not have been enough. On the other hand, the specifics of the effects of the intervention are probably not predicted from simple demand effects. Relatedly, a procedural difference existed in that only the distancing group received a strategy reminder during the second video block. While intended to facilitate real-time regulation, this prompt may have inadvertently encouraged retrospective reappraisal to align with the distancing narrative.

Second, employing a standard Kalman Filter approach to analyze the time-series of emotion ratings and video inputs, was based on several considerations. i) There are ample tools for the analysis and characterization of Kalman Filters readily available. ii) The Kalman Filter allows for observations to be noisy. This contrasts with typical analyses of EMA and similar time-series emotion ratings data, which do not allow for noise in the observations. This is important because the noise in ratings influences estimates of parameters as the rating error ‘persists’ in the modelled future. Nevertheless, one drawback of the Kalman Filter approach we employed was that Gaussian observation noise was assumed, which does not fully account for the bounded nature of the 0-100 emotional rating scales. However, reconstructed time-series were qualitatively close to the real time-series, and hence we do not believe that the fundamental conclusions here are likely to be affected by this. Future iterations could incorporate a truncated or censored observation model to better align the generative process with the constraints of behavioral rating scales.

Furthermore, online recruitment via Prolific Academic may introduce self-selection effects, and the sample may not be representative of the general population.

Finally, it is possible that eye gaze redirection away from emotionally charged regions in the videos could have acted as an intervening factor in emotion downregulation. The literature on emotion-specific oculomotor avoidance suggests that gaze patterns differ across emotion categories (Armstrong et al., 2014; Mulckhuyse, 2018; Pischek-Simpson et al., 2009; Kelly and Forsyth, 2009; Dalmaijer et al., 2021), which could contribute to differential effects on the input weight matrix for specific emotions. Future research could tackle that problem by tracking eye movement.

Conclusion

In conclusion, we used a novel methodological approach to characterize a key aspect of emotional states, namely how they can be controlled. We found that a brief distancing intervention can effectively regulate emotional experiences, reduce emotional variability and enhance the feeling of calmness. Moreover, distancing leads to qualitatively different changes in the dynamical structure of emotional states. First, it increases the intrinsic emotional stability, and second it reduces the impact of external inputs. Together, these alter the extent to which emotional states are externally controlled. Further computationally detailed characterization of emotional state dynamics and psychotherapeutic interventions may be useful on the path towards understanding differential and specific effects of different psychotherapeutic interventions.